---
title: "Explaining some `.assign()` idioms."
author: Ian D. Gow
date: 2026-02-07
date-format: "D MMMM YYYY"
number-sections: true
format:
  html:
    colorlinks: true
  pdf: 
    colorlinks: true
    geometry:
      - left=2cm
      - right=2cm
    papersize: a4
    mainfont: TeX Gyre Pagella
    mathfont: TeX Gyre Pagella Math
bibliography: papers.bib
csl: jfe.csl
engine: jupyter
jupyter: notes_py
---

Users of pandas will know about the `.assign` method of the pandas DataFrame.

```{python}
import pandas as pd
```

I borrow the toy example data frame from [pandas documentation](https://pandas.pydata.org/docs/reference/api/pandas.DataFrame.assign.html), which is `df` below.

```{python}
df = pd.DataFrame({"temp_c": [17.0, 25.0]}, 
                  index=["Portland", "Berkeley"])
df
```

The documentation tells us that the return value of the method is "a new DataFrame with the new columns in addition to all the existing columns."^[Existing columns that are re-assigned will be overwritten.]
We can see this here:

```{python}
df.assign(state=["WA", "CA"])
```

Note that the original DataFrame is unchanged.

```{python}
df
```
However, if we assign the value returned by `.assign()` to the variable associated with the original DataFrame, it will be practically as if we had modified that DataFrame. 

```{python}
df = df.assign(state=["WA", "CA"])
```

Looking at the [pandas documentation](https://pandas.pydata.org/docs/reference/api/pandas.DataFrame.assign.html) for the `.assign()` method, we see that its signature is `.assign(**kwargs)`.
This tells use that `.assign()` only takes keyword arguments.
In the example above, the keyword was `state`.

A more "traditional" approach might be the assignment to a Series as in the following.^[A good proxy for the "traditional" might be "Python for Data Analysis", written by the creator of pandas. The [third edition](https://wesmckinney.com/book/) of that book appears to contain not a single instance of `.assign()`.]

```{python}
df["state"] = ["WA", "CA"]
df
```

The documentation tells us that "if the values [passed to `.assign()`] are not callable, (e.g. a Series, scalar, or array), they are simply assigned."
This is precisely the behaviour we see above: the value is a list (`["WA", "CA"]`) and this is assigned to `temp_f` in the returned DataFrame.

But it is probably more common for the arguments to be a callable, such as in the following example, where an anonymous function (or lambda) is given as the value of the keyword `temp_f`.

```{python}
df.assign(temp_f=lambda x: x.temp_c * 9 / 5 + 32)
```

According to the documentation, "if the values are callable, they are computed on the DataFrame and assigned to the new columns", precisely the behaviour we see above.^[Another point made by the documentation is that "the callable must not change input DataFrame (though pandas doesn’t check it)."
This is fairly cryptic remark is perhaps alluding to the (seemingly) historical mutability of DataFrames.
I struggled, but was unable to contrive an example of actually violating this behaviour.]

Even this could be accomplished with the "traditional" approach:

```{python}
df["temp_f"] = df["temp_c"] * 9 / 5 + 32
df
```

So why depart from the "traditional" approach?
One reason is that `.assign()` fits nicely into the "method pipeline" approach that seems to have gained currency with data analysts in recent years.

```{python}
df = (
    pd.DataFrame({"temp_c": [17.0, 25.0]}, 
                  index=["Portland", "Berkeley"])
    .assign(temp_f=lambda x: x.temp_c * 9 / 5 + 32)
    .assign(temp_k=lambda x: (x.temp_f - 32) * 5 / 9 + 273.15)
)
df
```

```{python}
raw_data = {"city": ["Portland", "Berkeley"],
                   "temp_c": ["17", "25"],
                   "area_land": ["345.73 sq km", "27.02 sq km"],
                   "area_water": ["29.82 sq km", "18.71 sq km"],
                   "area_total": ["375.55 sq km", "45.73 sq km"],
                   "pop": ["652,503", "124,321"]}

df = pd.DataFrame(raw_data)
df
```

```{python}
df = (
    pd.DataFrame(raw_data)
    .assign(
      temp_c=lambda d: pd.to_numeric(d["temp_c"], errors="raise"),
      area_land=lambda d: pd.to_numeric(d["area_land"]
                          .str.replace(" sq km", "", regex=False), errors="raise"),
      area_water=lambda d: pd.to_numeric(d["area_water"]
                          .str.replace(" sq km", "", regex=False), errors="raise"),
      area_total=lambda d: pd.to_numeric(d["area_total"]
                          .str.replace(" sq km", "", regex=False), errors="raise"),
      pop=lambda d: pd.to_numeric(d["pop"]
                          .str.replace(",", "", regex=False), errors="raise"),
    )
)
df
```

Ew!
That's some ugly code.

```{python}
def parse_area(s: pd.Series) -> pd.Series:
    return pd.to_numeric(
        s.str.replace(" sq km", "", regex=False),
        errors="raise",
    )

def parse_pop(s: pd.Series) -> pd.Series:
    return pd.to_numeric(
        s.str.replace(",", "", regex=False),
        errors="raise",
    )                       
```

```{python}
df = (
    pd.DataFrame(raw_data)
    .assign(
        temp_c=lambda d: pd.to_numeric(d["temp_c"], errors="raise"),
        area_land=lambda d: parse_area(d["area_land"]),
        area_water=lambda d: parse_area(d["area_water"]),
        area_total=lambda d: parse_area(d["area_total"]),
        pop=lambda d: parse_pop(d["pop"]),
    )
)
df
```

```{python}
area_cols = ["area_land", "area_water", "area_total"]

df = (
    pd.DataFrame(raw_data)
    .assign(
        temp_c=lambda d: pd.to_numeric(d["temp_c"], errors="raise"),
        pop=lambda d: parse_pop(d["pop"]),
        **{c: (lambda d, c=c: parse_area(d[c]))
           for c in area_cols}
    )
)
df
```
  
```{r}
#| message: false
library(tidyverse)
```

```{r}
raw_data <- list(
  city = c("Portland", "Berkeley"),
  temp = c("17C", "25C"),
  area_land  = c("345.73 sq km", "27.02 sq km"),
  area_water = c("29.82 sq km", "18.71 sq km"),
  area_total = c("375.55 sq km", "45.73 sq km"),
  pop = c("652,503", "124,321")
)

df <- 
  raw_data |>
  as_tibble() |>
  mutate(across(-city, parse_number))

df
```

```{python}
def parse_number(s: pd.Series) -> pd.Series:
    return (
        s.astype("string")
         .str.replace(r"[^\d\.\-]+", "", regex=True)
         .pipe(pd.to_numeric, errors="coerce")
    )
```

```{python}
df = (
    pd.DataFrame(raw_data)
    .pipe(lambda d: d.assign(**{c: parse_number(d[c])
                                for c in d.columns if c != "city"})
    )
    .rename({"temp": "temp_c"})
    .assign(temp_f=lambda x: x.temp_c * 9 / 5 + 32)
    .assign(temp_k=lambda x: (x.temp_f - 32) * 5 / 9 + 273.15)
)

df
```
