---
title: "Aussie Supermarkets versus the World: Measuring Supermarket Profitability"
author: Professor Matt Pinnuck
date: 2024-03-28
date-format: "D MMMM YYYY"
format:
  pdf: 
    colorlinks: true
    geometry:
      - left=2.5cm
      - right=2.5cm
    papersize: a4
    mainfont: TeX Gyre Pagella
    mathfont: TeX Gyre Pagella Math
bibliography: papers.bib
---

```{r}
#| message: false
#| include: false
library(dplyr, warn.conflicts = FALSE)
library(DBI)
library(farr, warn.conflicts = FALSE)
library(dbplyr)
library(ggplot2)
library(stringr)
```

```{r}
#| include: false
gvkeys <- c("023667", "100346", "101173", "101971", "105104", "160381",
            "163896", "204459", "205876", "215406", "221102", "273498",
            "289488", "294564", "295761", "298073", "298181", "323355",
            "323984", "100131", "222398", "273961", "311342", "019579",
            "101465", "006502", "011259", "025645", "029028", "006771",
            "100894", "329498")

local_gvkeys <- c("100894", "329498")

db <- dbConnect(duckdb::duckdb())

g_funda <- load_parquet(db, table = "g_funda", schema = "comp")
funda <- load_parquet(db, table = "funda", schema = "comp")

vars <- c("gvkey", "datadate", "conm", "ceq",
          "fyear", "fic", "sale", "niadj", "nicon",
          "oiadp", "dltt", "dlc", "mib", "teq")

g_funda_mod <-
  g_funda |>
  filter(indfmt == "INDL", datafmt == "HIST_STD",
         consol == "C", popsrc == "I") |>
  select(any_of(vars)) |>
  rename(ni = nicon)

funda_mod <-
  funda |>
  filter(indfmt == "INDL", datafmt == "STD",
         consol == "C", popsrc == "D")  |>
  filter(fic != "MEX") |>
  select(any_of(vars)) |>
  rename(ni = niadj)

funda_all <-
  g_funda_mod |>
  union_all(funda_mod)

histsuper <-
  funda_all |>
  filter(gvkey %in% gvkeys) |>
  mutate(across(c(dltt, dlc, mib, teq), \(x) coalesce(x, 0))) |>
  mutate(ic = dltt + dlc + mib + teq) |>
  group_by(gvkey) |>
  window_order(datadate) |>
  mutate(lag_ceq = lag(ceq),
         lag_ic = lag(ic)) |>
  ungroup() |>
  filter(fyear >= 2021) |>
  mutate(ebitm = oiadp / sale,
         roic = oiadp / ((ic + lag_ic) / 2),
         roe = ni / (ceq + lag_ceq) / 2) |>
  collect()
```

```{r}
#| include: false
local_roic <-
  histsuper |>
  filter(gvkey %in% local_gvkeys) |>
  summarize(median(roic)) |>
  pull()

oecd_roic <-
  histsuper |>
  filter(!gvkey %in% local_gvkeys) |>
  summarize(median(roic)) |>
  pull()

local_roic_str <- round(local_roic * 100, 1)
oecd_roic_str <- round(oecd_roic * 100, 1)
```

```{r}
#| include: false
local_roe <-
  histsuper |>
  filter(gvkey %in% local_gvkeys) |>
  summarize(median(roe, na.rm = TRUE)) |>
  pull()

oecd_roe <-
  histsuper |>
  filter(!gvkey %in% local_gvkeys) |>
  summarize(median(roe, na.rm = TRUE)) |>
  pull()

local_roe_str <- round(local_roe * 100, 1)
oecd_roe_str <- round(oecd_roe * 100, 1)
```

```{r}
#| include: false
local_ebitm <-
  histsuper |>
  filter(gvkey %in% local_gvkeys) |>
  summarize(median(ebitm)) |>
  pull()

oecd_ebitm <-
  histsuper |>
  filter(!gvkey %in% local_gvkeys) |>
  summarize(median(ebitm)) |>
  pull()

local_ebitm_str <- round(local_ebitm * 100, 1)
oecd_ebitm_str <- round(oecd_ebitm * 100, 1)
```

# Summary
The perception of price gouging and realization of abnormal profits by Australian supermarkets has been a topic of debate. However, very little evidence has been provided as to what normal profits should be.
To provide a benchmark we analyse the profitability of Australian supermarkets against the largest supermarkets in OECD countries across the 3-year period 2021 to 2023.
We find no evidence that Australian supermarkets are making abnormal profits. 

Given that that supermarkets have large fixed and sunk costs, margins alone (the focus of much of the commentary) are insufficient to determine if profitability is excessive.
Rather an assessment is required of the: return on capital (ROC), the magnitude of fixed-cost investment and the margins necessary to recover fixed costs.
Across all metrics the evidence is consistent with Australian supermarkets being highly competitive, providing significant infrastructure benefits to Australian consumer (e.g., a large variety of offerings and on-line distribution platforms) and that shareholders are only making normal profits. 

# Return on Capital

The ROC of Woolworths Group and Coles of 13.2% and 14.6% is comparable to the median ROC for OECD supermarkets of `r oecd_roic_str`%.
The frequency distribution of ROC across the world shows that the ROC of Woolworths and Coles is clearly in the middle of the distribution of among the world’s supermarkets. 
See @fig-roic. 

```{r}
#| label: fig-roic
#| fig-cap: "Distribution of Return on Capital across OECD Supermarkets: 2021--2023"
#| echo: false
x_min <- round(min(histsuper$roic), 2)
x_max <- round(max(histsuper$roic), 2)
breaks <- seq(x_min, x_max, 0.02)

histsuper |>
  filter(!gvkey %in% local_gvkeys) |>
  ggplot(aes(x = roic)) +
  geom_density() +
  scale_x_continuous(breaks = breaks,
                     labels = scales::label_percent()) +
  geom_vline(xintercept = local_roic, color = "blue", linetype = "solid") +
  annotate("text", x = local_roic + 0.002, y = 7.5,
           label = str_c("Coles/Woolies = ", local_roic_str, "%"),
           color = "blue", angle = 360, hjust = 0) +
  geom_vline(xintercept = oecd_roic, color = "blue", linetype = "dashed") +
  annotate("text", x = oecd_roic - 0.002, y = 7.,
           label = str_c("OECD Median = ", oecd_roic_str, "%"),
           color = "blue", angle = 360, hjust = 1) +
  xlab("Return on Capital") + ylab("Frequency") +
  theme(axis.text = element_text(size = 7)) +
  xlim(c(x_min, x_max))
```

# Return on Equity

The frequency distribution of ROE across the world shows that the ROC of Woolworths and Coles is clearly in the middle of the distribution of among the world’s supermarkets. 
See @fig-roe. 

```{r}
#| label: fig-roe
#| fig-cap: "Distribution of Return on Equity across OECD Supermarkets: 2021--2023"
#| echo: false
x_min <- max(-0.06, max(min(histsuper$roe, na.rm = TRUE)))
x_max <- max(histsuper$roe, na.rm = TRUE)
breaks <- seq(round(x_min, 2), round(x_max, 2), 0.02)

histsuper |>
  filter(!gvkey %in% local_gvkeys, !is.na(roe), roe > x_min) |>
  ggplot(aes(x = roe)) +
  geom_density() +
  scale_x_continuous(breaks = breaks,
                     labels = scales::label_percent()) +
  geom_vline(xintercept = local_roe, color = "blue", linetype = "solid") +
  annotate("text", x = local_roe + 0.002, y = 7.5,
           label = str_c("Coles/Woolies = ", local_roe_str, "%"),
           color = "blue", angle = 360, hjust = 0) +
  geom_vline(xintercept = oecd_roe, color = "blue", linetype = "dashed") +
  annotate("text", x = oecd_roe - 0.002, y = 7.,
           label = str_c("OECD Median = ", oecd_roe_str, "%"),
           color = "blue", angle = 360, hjust = 1) +
  xlab("Return on Equity") + ylab("Frequency") +
  theme(axis.text = element_text(size = 7)) 
```

# Fixed Cost Investments and Quality of Service

One of the primary bases on which supermarkets compete for consumers’ business is investment o improve service quality and expansion of product variety (e.g. gluten-free, kimchi, organic food).
This requires investment in large new stores, store renewals, convenience stores, supply chain facilities and warehouses, on-line distribution technology and on-demand delivery, etc.
If there were no competition, there would be no need to make these risky fixed and usually sunk investments. 
The magnitude of capital investment provides an indicator of investment in fixed costs.
Capital investment to sales is 4.0% for Woolworths and 3.3% for Coles.
This is substantially greater than the median capital investment for OECD supermarkets of 2.6%.
As illustrated in the frequency distribution, the median CAPX investment by Woollies and Coles is greater than that of 75% of OECD supermarkets). 
See Figure 2. 
The magnitude of fixed investments by Australian supermarkets is strongly suggestive of a tightly contested market. 

# Margins

Consumers benefit substantially from these risky investments. 
However, there is no evidence that Woolworths Group and Coles are charging excessive margins to recover these fixed costs. 
The EBIT margins of Woolworths and Coles of 4.7% and 4.7% is comparable to the median EBIT margins for OECD supermarkets of 4.4%.
See @fig-ebitm.

```{r}
#| label: fig-ebitm
#| fig-cap: "Distribution of EBIT Margins across OECD Supermarkets: 2021--2023"
#| echo: false
breaks <- seq(round(min(histsuper$ebitm), 2), 
              round(max(histsuper$ebitm), 2), 
              0.01)
histsuper |>
  filter(!gvkey %in% local_gvkeys) |>
  ggplot(aes(x = ebitm)) +
  geom_density() +
  scale_x_continuous(breaks = breaks,
                     labels = scales::label_percent(accuracy = 1)) +
  geom_vline(xintercept = local_ebitm, color = "blue", linetype = "solid") +
  annotate("text", x = local_ebitm + 0.002, y = 2,
           label = str_c("Coles/Woolies = ", local_ebitm_str, "%"),
           color = "blue", angle = 360, hjust = 0) +
  geom_vline(xintercept = oecd_ebitm, color = "blue", linetype = "dashed") +
  annotate("text", x = oecd_ebitm - 0.002, y = 1,
           label = str_c("OECD Median = ", oecd_ebitm_str, "%"),
           color = "blue", angle = 360, hjust = 1) +
  xlab("ROC") + ylab("Frequency") +
  theme(axis.text = element_text(size = 7))
```

A further implication of the significant investment in fixed costs is that it gives rise to increased risk due to higher operating leverage and year-on-year changes in margins cannot be used in isolation as evidence of price gouging. 
With fixed costs, increased sales lead to higher EBIT margin due simply to fixed costs being spread over a greater volume.
The sales of both Coles and Woolworths in 2023 increased significantly implying that changes in margins are at least partly explained by the effects of operating leverage. 

# Summary
The magnitude of the fixed investments by Australian supermarkets is both strongly suggestive of a tightly contested market and requires sufficient margins to recover these fixed costs. The ROC and EBIT margins of Coles and Woolworth Group do not appear excessive relative to comparable OECD supermarkets. 

# Caveat
The reported figures in this analysis are based on Woolworths Group which includes the effects of the non-food divisions such as K-Mart. 
The self-reported EBIT margins and ROFE by the Woolworths Group for the Woolworths food division are significantly greater than the medians for the OECD benchmark supermarkets.

# DuPont analysis

As shareholders we may care about return on equity (*ROE*). 
Thus, understanding what drives *ROE* helps us better understand what drives stock value. 
The simple DuPont decomposition is an attempt to do this, but does not clearly separate the effects of operating vs. financing decisions.^[[Simple DuPont Decomposition: \\ $\textit{ROE}=\frac{NI}{E}=\frac{\textit{Sales}}{A}\times\frac{NI}{\textit{Sales}}\times\frac{A}{E}$.]
For example, net income includes (thus is confounded by) the effects of capital structure (e.g., interest expense), complicating the comparison between firms with different level of debt. 
Similarly, total assets includes both operating and financial assets.
The advanced DuPont decomposition introduces return on net operating assets (*RNOA*) as a central measure of operating performance to separate the effects of operating decisions from the effects of financial leverage on *ROE*. 

## Concepts and Definitions} {#sec-concepts}

To understand the advanced DuPont decomposition, it is useful to introduce the following concepts. 
These concepts arise from dividing the balance sheet into operating assets and liabilities and financial assets and liabilities. 
The following section will illustrate
how these concepts become useful. 

1. Net Operating Assets (*NOA*): Operating Working Capital + Net Operating Long-Term Assets
      
  - Operating Working Capital = Current Operating Assets - Current Operating Liabilities^[Current operating assets is current assets minus the amount of cash and marketable securities above what is considered "normal" or necessary to operate the business. 
Current operating liabilities is current liabilities minus short-term debt and the current portion of long-term debt.]
  - Net Operating Long-Term Assets = Total Long-Term Assets - Non-Interest-Bearing Long-Term Liabilities
          
    
2. Net Operating Profit After Taxes (*NOPAT*): Net Income + Net Interest Expenses After Tax^[The idea is to calculate the firm's profitability if it held no debt and has no interest expense.]
    
  - Net Interest Expense After Tax = (Interest Expense - Interest Income)$\times(1-\tau) = \textit{Int} \times(1-\tau)$
                $$
                    \textit{NOPAT} = NI + Int\times(1-\tau) 
                    = (\textit{EBIT}-Int)\times(1-\tau)+\textit{Int}\times(1-\tau) 
                    = \textit{EBIT}\times(1-\tau)
                $$
            
    
3. Return on *NOA* (*RNOA*): $= \textit{NOPAT}/\textit{NOA}$.^This measure is not affected by the amount of debt (relative to equity) because *NOA* and *NOPAT* are not affected by capital structure.]
    
    \item Spread: \textit{R\textit{NOA}--Net Cost of Debt}
            \begin{itemize}
                \item Net Cost of Debt = Net Interest Expense After Tax / Net Debt
                \item Net Debt = Financial Liabilities - Financial Assets\footnote{Financial liabilities is the sum of short-term debt, current portion of long-term debt, and interest-bearing long-term debt. Financial assets are the cash and marketable securities above what is considered ``normal'' or necessary to operate the business.}
            \end{itemize}    
    \item Net Financial Leverage: Net Debt/Equity
\end{enumerate}

## Derivation}\label{sec:Derivation}

\begin{enumerate}
    \item $A = L + E$ $^{[}$\footnote{We begin with the balance sheet equation}$^{]}$
    \bigskip
    \bigskip
    \item $OA + FA = OL + FL + E$ $^{[}$\footnote{We will divide assets and liabilities into operating and financial components}$^{]}$
    \bigskip
    \bigskip
    \item $\textit{NOA} = \textit{Net Debt} +E$ $^{[}$\footnote{We can re-express in terms of net operating assets and net debt}$^{]}$
    \bigskip
    \bigskip
    \item \begin{eqnarray*}
    \textit{ROE}&=&\frac{NI}{E}\\
       &=&\frac{\textit{EBIT}\times(1-\tau) - \textit{Int}\times(1-\tau)}{E}\\
       &=&\left[\frac{\textit{EBIT}\times(1-\tau)}{\textit{NOA}}\right] \times \frac{\textit{NOA}}{E} - \left[\frac{\textit{Int}\times(1-\tau)}{\textit{Net Debt}}\right] \times \frac{\textit{Net Debt}}{E} \\
       &=& \textit{RNOA} \times \frac{E+\textit{Net Debt}}{E} - \textit{Net Borrowing Cost} \times \frac{\textit{Net Debt}}{E} \\
       &=&\textit{RNOA}+ \left(\textit{RNOA}-\textit{Net Borrowing Cost}\right) \times \frac{\textit{Net Debt}}{E} \\ 
       &=&\textit{RNOA} + \textit{Spread}\times \textit{Net Fin Leverage}
    \end{eqnarray*}
        
\end{enumerate}

\bigskip
\section{Conceptual Notes}
\begin{enumerate}
    \item You can think of RNOA as the \textbf{ROA or ROE for an all-equity financed firm}. It takes out the effect of financing from both the numerator and the denominator of ROA, so we can focus solely on how well managers are operating the firm or deploying the company's net operating assets independent of financing decisions. 
    \item \textit{Spread}$\times$ \textit{Net Fin Leverage} is the \textbf{gain from financial leverage}. Another way to think of this is to re-express this gain as 
    \begin{equation*}
        \frac{Spread\times NetDebt}{Equity}.
    \end{equation*} 
    The numerator is the dollar return you generate to shareholders from taking on additional (net) debt: since you earn \textit{RNOA}$\times$\textit{NetDebt} dollars, but pay back an effective interest of \textit{NetInterestRate}$\times$\textit{NetDebt}, with the remainder representing the dollar amount that is attributable to equity holders. Dividing by total equity re-expresses this dollar amount as the percent return to equity that comes from taking out additional net debt. 
    \item The effective interest rate on net debt 
        \begin{equation*}
            \frac{(Interest Expense - Interest Income)\times(1-\tau)}{NetDebt}        
        \end{equation*}
        evaluates the effectiveness of the company's management of financial assets and liabilities. It punishes for holding high levels of financial assets (lower \textit{NetDebt}) that have low returns (lower \textit{Interest Income}), or for taking out small amount of loans at a high interest (higher \textit{InterestExpense} with lower \textit{NetDebt}). On the other hand, it rewards generating high level of financial returns with a low level of financial assets (high \textit{InterestIncome} and lower \textit{NetDebt}) and rewards low interest expenses especially at high levels of debt (lower \textit{InterestExpense} and higher \textit{NetDebt}). 
    \item Although \textit{Spread} may be positive at any point in time, the gain from leverage may not increase monotonically as we increase financial leverage. Above some efficient level of leverage, additional net debt may increase the cost of debt. Combined with diminishing returns to scale which could lower RNOA, one expects the spread to eventually decline as leverage increases. 
    \item What if the firm has net financial assets (i.e., $\textit{Net Debt} < 0)$? In this case, we can re-express the deomposition as follows: 
        \begin{eqnarray*}
                ROE &=& RNOA+(RNOA - Interest)\times\frac{NetDebt}{E} \\ 
                    &=& RNOA+(RNFA - RNOA)\times\frac{NFA}{E} 
         \end{eqnarray*}[\footnote{NFA = Net Financial Assets=-Net Debt. RNFA = Return on Net Financial Assets = (Interest Revenue - Interest Expense)(1-$\tau$) / Net Financial Assets = (Interest Expense-Interest Revenue)(1-$\tau$) / Net Debt  = Effective Interest Rate}] 
    
    In this case, equity is invested in operating and financial assets ($\textit{NOA} + \textit{NFA} = E$). Thus, if returns from financing assets exceed that of operating assets, then equity holders benefit when a greater proportion of the capital is invested in net financial assets. If RNOA exceeds RNFA, on the other hand, then investing a greater proportion of equity in financial assets is worse for equity holders. 
\end{enumerate}


## Implementation

1 Divide assets and liabilities on the balance sheet into operating and financial assets and liabilities.^[Technically, we can carefully divide cash into an operating or "normal" portion and an "excess" or financial portion. 
Similarly, when we look at balance sheet accounts, we can carefully separate components of accounts (e.g., "LT Investments") into portions that are operating and those that are financial. In practice, we often treat all cash as negative net debt, and all accounts other than debt, borrowings, or capital lease liabilities as operating.]
        
  - Net Operating Working Capital (CA - CL and exclude financial assets and liabilities)
  
  - Net Operating Long-Term Capital (LT Assets - LT Liabilities and exclude LT financial assets and LT financial liabilities)
Doing so allows us to re-express the standard balance sheet equation as follows:
    $$
    \begin{eqnarray*}
        A &=& L+E \\ 
        OA + FA &=& OL + FA + E \\
        NOA &=& ND + E
    \end{eqnarray*}
    $$
 -  Divide the income statement into an operating and financial income and expenses.^[Typically we just take interest expense and interest income as reported. However, there could be financial assets (e.g., short term or long term securities) that generate earnings reported elsewhere in the income statement which we can parse out. Alternatively, there could be interest revenues (e.g., from holding "normal" level of cash necessary for operations of the business) that could be considered operating earnings rather than financial.]
Note that how we divide the income statement should be consistent with the approach taken to divide the balance sheet. 
  
  - Compute NOPAT, RNOA, Spread, and Net Financial Leverage
    
  - Check that ROE is *exactly* equal to RNOA + Spread $\times$ Net Financial Leverage
\end{enumerate}
