---
title: "Aussie Supermarkets versus the World: Measuring Supermarket Profitability"
author: Professor Matt Pinnuck
date: 2024-03-28
date-format: "D MMMM YYYY"
format:
  pdf: 
    colorlinks: true
    geometry:
      - left=2.5cm
      - right=2.5cm
    papersize: a4
    mainfont: TeX Gyre Pagella
    mathfont: TeX Gyre Pagella Math
bibliography: papers.bib
---

```{r}
#| message: false
#| include: false
library(dplyr, warn.conflicts = FALSE)
library(DBI)
library(farr, warn.conflicts = FALSE)
library(dbplyr)
library(ggplot2)
library(stringr)
```

```{r}
#| include: false
gvkeys <- c("023667", "100346", "101173", "101971", "105104", "160381",
            "163896", "204459", "205876", "215406", "221102", "273498",
            "289488", "294564", "295761", "298073", "298181", "323355",
            "323984", "100131", "222398", "273961", "311342", "019579",
            "101465", "006502", "011259", "025645", "029028", "006771",
            "100894", "329498")

local_gvkeys <- c("100894", "329498")

db <- dbConnect(duckdb::duckdb())

g_funda <- load_parquet(db, table = "g_funda", schema = "comp")
funda <- load_parquet(db, table = "funda", schema = "comp")

vars <- c("gvkey", "datadate", "conm", "ceq",
          "fyear", "fic", "sale",
          "oiadp", "dltt", "dlc", "mib", "teq")

g_funda_mod <-
  g_funda |>
  filter(indfmt == "INDL", datafmt == "HIST_STD",
         consol == "C", popsrc == "I") |>
  select(all_of(vars))

funda_mod <-
  funda |>
  filter(indfmt == "INDL", datafmt == "STD",
         consol == "C", popsrc == "D")  |>
  filter(fic != "MEX") |>
  select(all_of(vars))

funda_all <-
  g_funda_mod |>
  union_all(funda_mod)

histsuper <-
  funda_all |>
  filter(gvkey %in% gvkeys) |>
  mutate(across(c(dltt, dlc, mib, teq), \(x) coalesce(x, 0))) |>
  mutate(ic = dltt + dlc + mib + teq) |>
  group_by(gvkey) |>
  window_order(datadate) |>
  mutate(lag_ic = lag(ic)) |>
  ungroup() |>
  filter(fyear >= 2021) |>
  mutate(ebitm = oiadp / sale,
         roic = oiadp / ((ic + lag_ic) / 2)) |>
  collect()
```

```{r}
#| include: false
local_roic <-
  histsuper |>
  filter(gvkey %in% local_gvkeys) |>
  summarize(median(roic)) |>
  pull()

oecd_roic <-
  histsuper |>
  filter(!gvkey %in% local_gvkeys) |>
  summarize(median(roic)) |>
  pull()

local_roic_str <- round(local_roic * 100, 1)
oecd_roic_str <- round(oecd_roic * 100, 1)
```

```{r}
#| include: false
local_ebitm <-
  histsuper |>
  filter(gvkey %in% local_gvkeys) |>
  summarize(median(ebitm)) |>
  pull()

oecd_ebitm <-
  histsuper |>
  filter(!gvkey %in% local_gvkeys) |>
  summarize(median(ebitm)) |>
  pull()

local_ebitm_str <- round(local_ebitm * 100, 1)
oecd_ebitm_str <- round(oecd_ebitm * 100, 1)
```

# Summary
The perception of price gouging and realization of abnormal profits by Australian supermarkets has been a topic of debate. However, very little evidence has been provided as to what normal profits should be.
To provide a benchmark we analyse the profitability of Australian supermarkets against the largest supermarkets in OECD countries across the 3-year period 2021 to 2023.
We find no evidence that Australian supermarkets are making abnormal profits. 

Given that that supermarkets have large fixed and sunk costs, margins alone (the focus of much of the commentary) are insufficient to determine if profitability is excessive.
Rather an assessment is required of the: return on capital (ROC), the magnitude of fixed-cost investment and the margins necessary to recover fixed costs.
Across all metrics the evidence is consistent with Australian supermarkets being highly competitive, providing significant infrastructure benefits to Australian consumer (e.g., a large variety of offerings and on-line distribution platforms) and that shareholders are only making normal profits. 

# Return on Capital

The ROC of Woolworths Group and Coles of 13.2% and 14.6% is comparable to the median ROC for OECD supermarkets of `r oecd_roic_str`%.
The frequency distribution of ROC across the world shows that the ROC of Woolworths and Coles is clearly in the middle of the distribution of among the world’s supermarkets. 
See @fig-roic. 

```{r}
#| label: fig-roic
#| fig-cap: "Distribution of Return on Capital across OECD Supermarkets: 2021--2023"
#| echo: false
breaks <- seq(-0.02, 0.30, 0.02)
histsuper |>
  filter(!gvkey %in% local_gvkeys) |>
  ggplot(aes(x = roic)) +
  geom_density() +
  scale_x_continuous(breaks = breaks) +
  geom_vline(xintercept = local_roic, color = "blue", linetype = "solid") +
  annotate("text", x = local_roic + 0.002, y = 7.5,
           label = str_c("Coles/Woolies = ", local_roic_str, "%"),
           color = "blue", angle = 360, hjust = 0) +
  geom_vline(xintercept = oecd_roic, color = "blue", linetype = "dashed") +
  annotate("text", x = oecd_roic - 0.002, y = 7.,
           label = str_c("OECD Median = ", oecd_roic_str, "%"),
           color = "blue", angle = 360, hjust = 1) +
  xlab("EBIT Margin") + ylab("Frequency")
```

# Fixed Cost Investments and Quality of Service

One of the primary bases on which supermarkets compete for consumers’ business is investment o improve service quality and expansion of product variety (e.g. gluten-free, kimchi, organic food).
This requires investment in large new stores, store renewals, convenience stores, supply chain facilities and warehouses, on-line distribution technology and on-demand delivery, etc.
If there were no competition, there would be no need to make these risky fixed and usually sunk investments. 
The magnitude of capital investment provides an indicator of investment in fixed costs.
Capital investment to sales is 4.0% for Woolworths and 3.3% for Coles.
This is substantially greater than the median capital investment for OECD supermarkets of 2.6%.
As illustrated in the frequency distribution, the median CAPX investment by Woollies and Coles is greater than that of 75% of OECD supermarkets). 
See Figure 2. 
The magnitude of fixed investments by Australian supermarkets is strongly suggestive of a tightly contested market. 

# Margins

Consumers benefit substantially from these risky investments. 
However, there is no evidence that Woolworths Group and Coles are charging excessive margins to recover these fixed costs. 
The EBIT margins of Woolworths and Coles of 4.7% and 4.7% is comparable to the median EBIT margins for OECD supermarkets of 4.4%.
See @fig-ebitm.

```{r}
#| label: fig-ebitm
#| fig-cap: "Distribution of EBIT Margins across OECD Supermarkets: 2021--2023"
#| echo: false
breaks <- seq(-0.005, 0.09, 0.01)
histsuper |>
  filter(!gvkey %in% local_gvkeys) |>
  ggplot(aes(x = ebitm)) +
  geom_density() +
  scale_x_continuous(breaks = breaks) +
  geom_vline(xintercept = local_ebitm, color = "blue", linetype = "solid") +
  annotate("text", x = local_ebitm + 0.002, y = 2,
           label = str_c("Coles/Woolies = ", local_ebitm_str, "%"),
           color = "blue", angle = 360, hjust = 0) +
  geom_vline(xintercept = oecd_ebitm, color = "blue", linetype = "dashed") +
  annotate("text", x = oecd_ebitm - 0.002, y = 1,
           label = str_c("OECD Median = ", oecd_ebitm_str, "%"),
           color = "blue", angle = 360, hjust = 1) +
  xlab("ROC") + ylab("Frequency")
```

A further implication of the significant investment in fixed costs is that it gives rise to increased risk due to higher operating leverage and year-on-year changes in margins cannot be used in isolation as evidence of price gouging. 
With fixed costs, increased sales lead to higher EBIT margin due simply to fixed costs being spread over a greater volume.
The sales of both Coles and Woolworths in 2023 increased significantly implying that changes in margins are at least partly explained by the effects of operating leverage. 

# Summary
The magnitude of the fixed investments by Australian supermarkets is both strongly suggestive of a tightly contested market and requires sufficient margins to recover these fixed costs. The ROC and EBIT margins of Coles and Woolworth Group do not appear excessive relative to comparable OECD supermarkets. 

# Caveat
The reported figures in this analysis are based on Woolworths Group which includes the effects of the non-food divisions such as K-Mart. 
The self-reported EBIT margins and ROFE by the Woolworths Group for the Woolworths food division are significantly greater than the medians for the OECD benchmark supermarkets. 
