---
title: "A quick look at City of Melbourne bike data"
author: Ian D. Gow
date: 2024-09-19
date-format: "D MMMM YYYY"
format:
  html:
    colorlinks: true
  pdf: 
    classoption: [landscape]
    colorlinks: true
    geometry:
      - left=2.5cm
      - right=2.5cm
    papersize: a4
    mainfont: TeX Gyre Pagella
    mathfont: TeX Gyre Pagella Math
bibliography: papers.bib
---

In writing this note, I use the packages listed below.^[Execute `install.packages(c("tidyverse", duckdb", "tinytable"))` within R to install all the packages you need to run the code in this note.]
This note was written using [Quarto](https://quarto.org) and compiled with [RStudio](https://posit.co/products/open-source/rstudio/), an integrated development environment (IDE) for working with R.
The source code for this note is available [here](https://raw.githubusercontent.com/iangow/notes/main/bikes.qmd) and the latest version of this PDF is [here](https://raw.githubusercontent.com/iangow/notes/main/bikes.pdf).^[Some parts of the source code are ugly as I wrangled hurriedly with the output from SQL and LaTeX tables.]

```{r}
#| message: false
library(tidyverse)
library(duckdb)
library(tinytable)
```

The following code downloads the data and unzips the single file therein.
It is cached (`cache: true` in chunk options) to save time with repeated runs of the code.

```{r}
#| cache: true
t <- tempfile(fileext = ".zip")
url <- str_c("https://opendatasoft-s3.s3.amazonaws.com/",
             "downloads/archive/74id-aqj9.zip")
download.file(url, t)
unzip(t)
```
I create a database connection and load the `icu` extension, which contains time-zone information.^[You may need to run `INSTALL icu` before `LOAD icu` depending on your DuckDB installation.]

```{r}
#| results: false
db <- dbConnect(duckdb::duckdb(), timezone_out = "Australia/Melbourne")

dbExecute(db, "LOAD icu")
```

The following SQL creates `bikes_raw`, which is fairly unprocessed data.
Only `RUNDATE` is given a type, and this is `TIMESTAMP` because there is no time zone information in the data.

The following function is created in R, but generates SQL.
The [documentation](https://duckdb.org/docs/sql/functions/timestamptz.html) for `make_timestamptz()` says that it returns "the `TIMESTAMP WITH TIME ZONE` for the given µs since the epoch."
But it seems the data we have are in milliseconds, not microseconds, so we need to multiply by 1000.

```{r}
epoch_to_ts <- function(x) {
  x <- rlang::as_name(rlang::enquo(x))
  dplyr::sql(stringr::str_c("make_timestamptz(", x, " * 1000)"))
}
```

The following code converts `rundate` to `TIMESTAMPTZ` assuming the original data are Melbourne times.
It also converts `lastcommwithserver`, `latestupdatetime`, and `installdate` to `TIMESTAMPTZ`.
Note that attention needs to be paid to time zones, because the **epoch** is [defined](https://www.postgresql.org/docs/current/functions-datetime.html) as "the number of seconds since `1970-01-01 00:00:00 UTC`", which would be a different point in time from `1970-01-01 00:00:00` in Melbourne time.

```{r}
bikes <-
  tbl(db, "read_csv('74id-aqj9.csv',
                timestampformat='%Y%m%d%H%M%S',
                types={'RUNDATE': 'TIMESTAMP'})") |>
  rename_with(str_to_lower) |>
  select(-installed, -temporary, -removaldate) |>
  mutate(rundate = timezone("Australia/Melbourne", rundate),
         lastcommwithserver = !!epoch_to_ts(lastcommwithserver),
         latestupdatetime = !!epoch_to_ts(latestupdatetime),
         installdate = !!epoch_to_ts(installdate)) |>
  compute(name = "bikes", overwrite = TRUE) 
```

```{r}
dbExecute(db, "COPY bikes TO 'bikes.parquet'")
```

```{r}
gwalkr(bikes |> collect(), kernelComputation = TRUE)
```

The following SQL produces some information on the contents of `bikes` that is shown in @tbl-bikes.

```{sql}
--| connection: db
--| max.print: -1
--| output.var: results
SELECT column_name, column_type, max, null_percentage
FROM (SUMMARIZE bikes);
```

```{r}
#| label: tbl-bikes
#| echo: false
#| eval: true
#| tbl-cap: Information on processed data (`bikes`)
results |> tt() |> format_tt(escape = TRUE)
```

```{r}
#| label: tbl-time-sample
#| tbl-cap: Sample of date-time variables
#| render: !expr function(x, ...) mutate(x, across(everything(), as.character)) |> tt() |> knitr::knit_print()
bikes |> 
  select(lastcommwithserver, latestupdatetime, rundate, installdate) |>
  collect(n = 10)
```

In making @fig-runtime, I convert the date component of `runtime` to the same date (`2017-01-01`).
This facilitates plotting in R, as R has no native "time" type and thus things are easier using date-times.
Unfortunately, it seems that all the timestamps in `bikes` are boring back-end times produced by systems, so there is nothing special about the distibution of these times.
More interest plots might come from looking at when bikes are checked out and in (only net checkouts seem to be available) assuming that the data are sufficiently frequent.

```{r}
#| label: fig-runtime
#| fig-cap: Distribution of times in `runtime`
bikes |>
  mutate(runtime = make_timestamptz(2017L, 1L, 1L, 
                                    hour(rundate), minute(rundate), 
                                    second(rundate))) |>
  ggplot(aes(runtime)) +
  geom_histogram(binwidth = 60 * 60) +
  scale_x_datetime(date_breaks = "1 hour", date_labels = "%H")
```
