---
title: "From subqueries to CTEs"
author: 
  - name: Ian D. Gow
    orcid: 0000-0002-6243-8409
    email: iandgow@gmail.com
abstract: "Take a query with subqueries and turn it into one with CTEs."
date: today
date-format: "D MMMM YYYY"
bibliography: papers.bib
title-block-style: plain
format: 
    pdf:
        toc: false
        number-sections: true
        colorlinks: true
        mainfont: TeX Gyre Pagella
        sansfont: TeX Gyre Pagella
    html:
        theme: cosmo
---

In this brief note, I use a query from @tanimura2021sql to illustrate how one can re-write a query that makes use of subqueries as a query using common table expressions (CTEs).
I then show how a query written with CTEs is easily translated into a query using `dbplyr`.
Finally, I do the analysis again from scratch, but using `dbplyr` expressions.
In this analysis I find that the SQL query @tanimura2021sql does not quite meet the verbal specification it was intended to meet, while the `dbplyr` query does.
I conjecture that the "building blocks" approach to SQL facilitated by `dbplyr` may enhance the accuracy of queries for many users.

## The data

The example query I study below comes from Chapter 4 of @tanimura2021sql.
The following packages will be used and should be installed.

```{r}
#| warning: false
library(DBI)
library(tidyverse)
library(dbplyr)
library(duckdb)
```

We first get the data, which requires an internet connection.

```{r}
#| warning: false
#| eval: true 
#| cache: true
url <- paste0("https://raw.githubusercontent.com/cathytanimura/",
              "sql_book/master/Chapter%204%3A%20Cohorts/")
legislators_df <- read_csv(paste0(url, "legislators.csv"),
                           show_col_types = FALSE)
legislators_terms_df <- read_csv(paste0(url, "legislators_terms.csv"),
                                 show_col_types = FALSE)
```

```{r}
#| warning: false
#| include: false
#| eval: false
legislators_df <- read_csv("data/legislators.csv.gz",
                           show_col_types = FALSE)
legislators_terms_df <- read_csv("data/legislators_terms.csv.gz",
                                 show_col_types = FALSE)
```

```{r}
#| include: false
library(knitr)
knit_print.lazy_tbl <- function(x, ...) {
  knitr::knit_print(knitr::kable(collect(x, n = 5),
                                 digits = 4))
}

registerS3method("knit_print", "lazy_tbl", knit_print.lazy_tbl)
```     

```{r}
#| include: false
knit_print.data.frame <- function(x, ...) {
  knitr::knit_print(kable(x, digits = 4))
}

registerS3method("knit_print", "data.frame", knit_print.data.frame)
```

With the data in hand, we copy the two data tables to an in-memory DuckDB database.

```{r}
db <- dbConnect(duckdb::duckdb())
legislators <- copy_to(db, legislators_df, "legislators")
legislators_terms <- copy_to(db, legislators_terms_df, 
                             "legislators_terms")
```

## The query

Chapter 4 of @tanimura2021sql contains the following query.

```{sql}
#| connection: db
SELECT aa.cohort_century::int as cohort_century
,round(bb.rep_and_sen_5_yrs * 1.0 / aa.reps,4) as pct_5_yrs
,round(bb.rep_and_sen_10_yrs * 1.0 / aa.reps,4) as pct_10_yrs
,round(bb.rep_and_sen_15_yrs * 1.0 / aa.reps,4) as pct_15_yrs
FROM
(
        SELECT date_part('century',a.first_term) as cohort_century
        ,count(id_bioguide) as reps
        FROM
        (
                SELECT id_bioguide, min(term_start) as first_term
                FROM legislators_terms
                WHERE term_type = 'rep'
                GROUP BY 1
        ) a
        WHERE first_term <= '2009-12-31'
        GROUP BY 1
) aa
LEFT JOIN
(
        SELECT date_part('century',b.first_term) as cohort_century
        ,count(distinct case when age(c.term_start, b.first_term) <= interval '5 years' 
            then b.id_bioguide end) as rep_and_sen_5_yrs
        ,count(distinct case when age(c.term_start, b.first_term) <= interval '10 years' 
            then b.id_bioguide end) as rep_and_sen_10_yrs
        ,count(distinct case when age(c.term_start, b.first_term) <= interval '15 years' 
            then b.id_bioguide end) as rep_and_sen_15_yrs
        FROM
        (
                SELECT id_bioguide, min(term_start) as first_term
                FROM legislators_terms
                WHERE term_type = 'rep'
                GROUP BY 1
        ) b
        JOIN legislators_terms c on b.id_bioguide = c.id_bioguide
        and c.term_type = 'sen' and c.term_start > b.first_term
        GROUP BY 1
) bb on aa.cohort_century = bb.cohort_century
ORDER BY 1;
```

This query is quite complex.
Can we simplify it using CTEs?

First, notice that the subqueries labelled `a` and `b` are identical.
Let's clean that up.
We'll put `a` as a CTE at the beginning of the query (after `WITH`) and refer to that both where we currently have it, and also in place of `b`.
All references to `b` are changed to references to `a`.
We can then run the query to check that we are still getting the same results.

```{sql}
#| connection: db
WITH a AS (
    SELECT id_bioguide, min(term_start) as first_term
    FROM legislators_terms
    WHERE term_type = 'rep'
    GROUP BY 1) 

SELECT aa.cohort_century::int as cohort_century
,round(bb.rep_and_sen_5_yrs * 1.0 / aa.reps,4) as pct_5_yrs
,round(bb.rep_and_sen_10_yrs * 1.0 / aa.reps,4) as pct_10_yrs
,round(bb.rep_and_sen_15_yrs * 1.0 / aa.reps,4) as pct_15_yrs
FROM
(
        SELECT date_part('century',a.first_term) as cohort_century
        ,count(id_bioguide) as reps
        FROM a
        WHERE first_term <= '2009-12-31'
        GROUP BY 1
) aa
LEFT JOIN
(
        SELECT date_part('century', a.first_term) as cohort_century
        ,count(distinct case when age(c.term_start, a.first_term) <= interval '5 years' 
            then a.id_bioguide end) as rep_and_sen_5_yrs
        ,count(distinct case when age(c.term_start, a.first_term) <= interval '10 years' 
            then a.id_bioguide end) as rep_and_sen_10_yrs
        ,count(distinct case when age(c.term_start, a.first_term) <= interval '15 years' 
            then a.id_bioguide end) as rep_and_sen_15_yrs
        FROM a
        JOIN legislators_terms c on a.id_bioguide = c.id_bioguide
        and c.term_type = 'sen' and c.term_start > a.first_term
        GROUP BY 1
) bb on aa.cohort_century = bb.cohort_century
ORDER BY 1;
```

Next, let's do the same for `aa` and `bb`.
Note that there are commas after the CTEs defining `a` and `aa`, but not after `bb`, as it is the last CTE before the body of the query.
Again we run the query to check that we still have the same results.

```{sql}
#| connection: db
WITH 

a AS (
    SELECT id_bioguide, min(term_start) as first_term
    FROM legislators_terms
    WHERE term_type = 'rep'
    GROUP BY 1),
    
aa AS (
    SELECT date_part('century',a.first_term) as cohort_century,
        count(id_bioguide) as reps
    FROM a
    WHERE first_term <= '2009-12-31'
    GROUP BY 1),

bb AS (
    SELECT date_part('century', a.first_term) as cohort_century,
        count(distinct case when age(c.term_start, a.first_term) <= interval '5 years' 
            then a.id_bioguide end) as rep_and_sen_5_yrs,
        count(distinct case when age(c.term_start, a.first_term) <= interval '10 years' 
            then a.id_bioguide end) as rep_and_sen_10_yrs,
        count(distinct case when age(c.term_start, a.first_term) <= interval '15 years' 
            then a.id_bioguide end) as rep_and_sen_15_yrs
    FROM a
    JOIN legislators_terms c on a.id_bioguide = c.id_bioguide
    and c.term_type = 'sen' and c.term_start > a.first_term
    GROUP BY 1) 

SELECT aa.cohort_century::int as cohort_century
,round(bb.rep_and_sen_5_yrs * 1.0 / aa.reps,4) as pct_5_yrs
,round(bb.rep_and_sen_10_yrs * 1.0 / aa.reps,4) as pct_10_yrs
,round(bb.rep_and_sen_15_yrs * 1.0 / aa.reps,4) as pct_15_yrs
FROM aa
LEFT JOIN bb on aa.cohort_century = bb.cohort_century
ORDER BY 1;
```

Now we can delete the `aa` and `bb` labels and references to them.
We can also tidy up the main query a little, including using the more elegant (in my view) `USING` syntax.

Also notice that we have `age(c.term_start, a.first_term)` three times in the query above.
We can split `bb` into two, as we do below (now `bbb` followed by `bb`).

```{sql}
#| connection: db
WITH 

a AS (
    SELECT id_bioguide, min(term_start) as first_term
    FROM legislators_terms
    WHERE term_type = 'rep'
    GROUP BY 1),
    
aa AS (
    SELECT date_part('century', first_term) as cohort_century,
        count(id_bioguide) as reps
    FROM a
    WHERE first_term <= '2009-12-31'
    GROUP BY 1),
    
bbb AS (
    SELECT date_part('century', a.first_term) as cohort_century,
        a.id_bioguide,
        age(c.term_start, a.first_term) AS age
    FROM a
    JOIN legislators_terms c 
    ON a.id_bioguide = c.id_bioguide
    AND c.term_type = 'sen' AND c.term_start > a.first_term), 

bb AS (
    SELECT cohort_century,
        count(distinct case when age <= interval '5 years' 
            then id_bioguide end) as rep_and_sen_5_yrs,
        count(distinct case when age <= interval '10 years'
            then id_bioguide end) as rep_and_sen_10_yrs,
        count(distinct case when age <= interval '15 years' 
            then id_bioguide end) as rep_and_sen_15_yrs
    FROM bbb
    GROUP BY 1) 

SELECT cohort_century as cohort_century,
    round(rep_and_sen_5_yrs * 1.0 / reps, 4) AS pct_5_yrs,
    round(rep_and_sen_10_yrs * 1.0 / reps, 4) AS pct_10_yrs,
    round(rep_and_sen_15_yrs * 1.0 / reps, 4) AS pct_15_yrs
FROM aa
LEFT JOIN bb 
USING (cohort_century)
ORDER BY 1;
```

We are still left with the meaningless labels (e.g., `a`, `bb`).
We can give the CTEs more meaningful labels.
We also clean up `ages` a little (e.g., `USING`) and move `cohort_century` to `first_rep_terms`.
Again we check that results are the same.

```{sql}
#| connection: db
WITH 

first_rep_terms AS (
    SELECT id_bioguide,
        min(term_start) AS first_term,
        date_part('century', min(term_start)) AS cohort_century,
    FROM legislators_terms
    WHERE term_type = 'rep'
    GROUP BY 1),

cohort_sizes AS (
    SELECT cohort_century, count(id_bioguide) AS reps
    FROM first_rep_terms
    WHERE first_term <= '2009-12-31'
    GROUP BY 1),
    
ages AS (
    SELECT cohort_century,
        id_bioguide,
        age(term_start, first_term) AS age
    FROM first_rep_terms
    JOIN legislators_terms
    USING (id_bioguide)
    WHERE term_type = 'sen' AND term_start > first_term), 

age_cuts AS (
    SELECT cohort_century,
        count(distinct case when age <= interval '5 years' 
            then id_bioguide end) as rep_and_sen_5_yrs,
        count(distinct case when age <= interval '10 years'
            then id_bioguide end) as rep_and_sen_10_yrs,
        count(distinct case when age <= interval '15 years' 
            then id_bioguide end) as rep_and_sen_15_yrs
    FROM ages
    GROUP BY 1) 

SELECT cohort_century AS cohort_century,
    round(rep_and_sen_5_yrs * 1.0 / reps, 4) AS pct_5_yrs,
    round(rep_and_sen_10_yrs * 1.0 / reps, 4) AS pct_10_yrs,
    round(rep_and_sen_15_yrs * 1.0 / reps, 4) AS pct_15_yrs
FROM cohort_sizes
LEFT JOIN age_cuts 
USING (cohort_century)
ORDER BY 1;
```

Finally, let's put the "main" query in a CTE too.

```{sql}
#| connection: db
WITH 

first_rep_terms AS (
    SELECT id_bioguide,
        min(term_start) AS first_term,
        date_part('century', min(term_start)) AS cohort_century,
    FROM legislators_terms
    WHERE term_type = 'rep'
    GROUP BY 1),

cohort_sizes AS (
    SELECT cohort_century, count(id_bioguide) AS reps
    FROM first_rep_terms
    WHERE first_term <= '2009-12-31'
    GROUP BY 1),
    
ages AS (
    SELECT cohort_century,
        id_bioguide,
        age(term_start, first_term) AS age
    FROM first_rep_terms
    JOIN legislators_terms
    USING (id_bioguide)
    WHERE term_type = 'sen' AND term_start > first_term), 

age_cuts AS (
    SELECT cohort_century,
        count(distinct case when age <= interval '5 years' 
            then id_bioguide end) as rep_and_sen_5_yrs,
        count(distinct case when age <= interval '10 years'
            then id_bioguide end) as rep_and_sen_10_yrs,
        count(distinct case when age <= interval '15 years' 
            then id_bioguide end) as rep_and_sen_15_yrs
    FROM ages
    GROUP BY 1),

cohort_retention AS (
  SELECT cohort_century::int as cohort_century,
    round(rep_and_sen_5_yrs * 1.0 / reps, 4) AS pct_5_yrs,
    round(rep_and_sen_10_yrs * 1.0 / reps, 4) AS pct_10_yrs,
    round(rep_and_sen_15_yrs * 1.0 / reps, 4) AS pct_15_yrs
  FROM cohort_sizes
  LEFT JOIN age_cuts 
  USING (cohort_century))
  
SELECT *
FROM cohort_retention
ORDER BY cohort_century;
```

What's the value of this last step?
Well it means we can easily edit the query to debug the CTEs that are used.
For example, we could put the following at then end of the query about to look into `first_rep_terms`.

```{sql}
#| connection: db
#| eval: false
SELECT *
FROM first_rep_terms;
```

## Translating to dbplyr

Now that we have a query based on CTEs, it is *much* easier to translate to `dbplyr`.

```{r}
first_rep_terms <-
    legislators_terms %>%
    filter(term_type == 'rep') %>%
    group_by(id_bioguide) %>%
    summarize(first_term = min(term_start, na.rm = TRUE),
              .groups = "drop") %>%
    mutate(cohort_century = century(first_term))
```

```{r}
cohort_sizes <-
    first_rep_terms %>%
    filter(first_term <= '2009-12-31') %>%
    group_by(cohort_century) %>%
    summarize(reps = n())
```

```{r} 
ages <-
    first_rep_terms %>%
    inner_join(legislators_terms, by = "id_bioguide") %>%
    filter(term_type == 'sen', term_start > first_term) %>%
    mutate(age = age(term_start, first_term)) %>%
    select(cohort_century, id_bioguide, age)
```

```{r}
age_cuts <-
    ages %>%
    mutate(id_05 = if_else(age <= years(5), id_bioguide, NA),
           id_10 = if_else(age <= years(10), id_bioguide, NA),
           id_15 = if_else(age <= years(15), id_bioguide, NA)) %>%
    group_by(cohort_century) %>%
    summarize(rep_and_sen_5_yrs = n_distinct(id_05),
              rep_and_sen_10_yrs = n_distinct(id_10),
              rep_and_sen_15_yrs = n_distinct(id_15))
```

```{r}
cohort_sizes %>%
    left_join(age_cuts, by = "cohort_century") %>%
    mutate(pct_5_yrs = rep_and_sen_5_yrs * 1.0 / reps,
           pct_10_yrs = rep_and_sen_10_yrs * 1.0 / reps,
           pct_15_yrs = rep_and_sen_15_yrs * 1.0 / reps) %>%
    mutate(across(starts_with("pct_"), ~ round(., 4))) %>%
    select(cohort_century, starts_with("pct_")) %>%
    arrange(cohort_century) %>%
    collect() 
```

## Doing it again in dplyr

Now let's do it again more or less from scratch using `dbplyr`.
In this version, I will build up block by block in a way that is easier (at least for me) to reason about.

For convenience, I reproduce the code we used to create `first_rep_terms` above.

```{r}
first_rep_terms <-
    legislators_terms %>%
    filter(term_type == 'rep') %>%
    group_by(id_bioguide) %>%
    summarize(first_rep_term = min(term_start, na.rm = TRUE),
              .groups = "drop")
```

Next, I produce an equivalent table for senate terms.

```{r}
first_sen_terms <-
    legislators_terms %>%
    filter(term_type == 'sen') %>%
    group_by(id_bioguide) %>%
    summarize(first_sen_term = min(term_start, na.rm = TRUE),
              .groups = "drop") 
```

We are interested in [legislators] "start as representatives" as our cohort.
We want to exclude those who start as senators and (to match the query in the book), we also want to exclude those who first term is after `'2009-12-31'`.

```{r}
cohort_members <-
  first_rep_terms %>%
  left_join(first_sen_terms, by = "id_bioguide") %>%
  filter(first_rep_term < first_sen_term | is.na(first_sen_term)) %>%
  select(id_bioguide, first_rep_term) %>%
  mutate(cohort_century = century(first_rep_term)) %>%
  filter(first_rep_term <= '2009-12-31')
```

We can now calculate the sizes of the cohorts that we have formed.^[Here I am using the term "cohort" in a way that I object to below. Seems like another term is needed for precision here.]

```{r}
cohort_sizes <-
  cohort_members %>%
  group_by(cohort_century) %>%
  summarize(reps = n(), .groups = "drop")
```

Now let's turn to the "go on to become senators" part of our remit.
This is a simple `INNER JOIN` with an inequality `first_rep_term < first_sen_term` condition.
We can calculate the `gap` using the SQL function `age()`.

```{r}
rep_then_sen <-
  cohort_members %>%
  inner_join(first_sen_terms, 
            join_by(id_bioguide, 
                    first_rep_term < first_sen_term)) %>%
  mutate(gap = age(first_sen_term, first_rep_term))
```

It is easy to check that `id_bioguide` is a valid key for `rep_then_sen`.
This fact makes it easy to building up the curve of subsequent senate terms using a window function.
Within each `cohort_century` we sum up the number of rows leading up to each value of `age`.
This is equivalent to `count(DISTINCT id_bioguide) OVER (PARTITION BY cohort_century ORDER BY gap)`, but we do not need the `DISTINCT` here because each value of `id_bioguide` is unique here.

Note that there will be ties in terms of `age` here, and we deal wtih these using `max()`.
That is if we had 432 representatives with `gap` of less than 5 years and 6 with `gap` of exactly 5 years, we want to step from 432 to 438 immediately.

```{r}
rep_then_sen_gaps <-
  rep_then_sen %>%
  group_by(cohort_century) %>%
  window_order(gap) %>%
  mutate(cum_ids = cumsum(1)) %>%
  group_by(cohort_century, gap) %>%
  mutate(cum_ids = max(cum_ids, na.rm = TRUE)) %>%
  ungroup()
```

I can now combine the "start as representatives" table (`cohort_sizes`) with the "go on to become senators" table (`rep_then_sen_gaps`) to calculate the percentages by `cohort_century`.

```{r}
pct_rep_then_sen <-
  rep_then_sen_gaps %>%
  inner_join(cohort_sizes, by = "cohort_century") %>%
  mutate(pct = cum_ids/reps) 
```

Rather than writing complicated `CASE` statements, I can make a little table in R with the three cutoff values and sent that to DuckDB and turn the rows into intervals.

```{r}
gap_cutoffs <-
  tibble(cutoff = c(5, 10, 15)) %>%
  copy_to(db, ., name = "gap_cutoffs",
          overwrite = TRUE) %>%
  mutate(cutoff = years(cutoff))
```

Now I `CROSS JOIN` `pct_rep_then_sen` and `gap_cutoffs` and calculate the `pct` values for each before using `pivot_wider` to rearrange the table to match what is shown in @tanimura2021sql.

```{r}
pct_rep_then_sen %>%
  cross_join(gap_cutoffs) %>%
  filter(gap <= cutoff) %>%
  group_by(cohort_century, cutoff) %>%
  summarize(pct = max(pct, na.rm = TRUE),
            .groups = "drop") %>%
  mutate(cutoff = year(cutoff)) %>%
  pivot_wider(names_from = "cutoff",
              names_prefix = "pct_",
              values_from = "pct") %>%
  arrange(cohort_century) %>%
  collect()
```

But we now see different numbers.
What has happened?

It turns out that there are two situations I addressed in the code that produced the table just above.
First, I only included in our cohort members of Congress who *started* as representatives.
The original SQL query included cohort members of Congress who had served as representatives, but who had previously served as senators (so had not *started as* representatives).
Second, I was consistent in application of the `first_rep_term <= '2009-12-31'` criterion throughout.
The original SQL query imposed this requirement in calculating `cohort_sizes`, but not in constructing the cohorts themselves.

I would argue that the answer I produced better matches the original question from @tanimura2021sql: 
"What share of [legislators] start as representatives and go on to become senators? 
(Some senators later become representatives, but that is much less common.)
Since relatively few make this transition, we'll cohort legislators by the century in which they first became a representative."^[Minor edits here. I am not a fan of this use of the term "cohort" as the cohorts are formed based on `rep_first_date` and only aggregation of statistics happens by `cohort_century`.]

Because we have access to the underlying tables as lazy data frames in R (effectively subqueries available for use to inspect), it is easy to dig into the steps of the two queries to check our reasoning.

```{r}
last_sen_terms <-
    legislators_terms %>%
    filter(term_type == 'sen') %>%
    group_by(id_bioguide) %>%
    summarize(last_sen_term = max(term_start, na.rm = TRUE),
              .groups = "drop") 

first_rep_terms %>%
  anti_join(cohort_members, by = "id_bioguide") %>%
  inner_join(first_sen_terms, by = "id_bioguide") %>%
  inner_join(last_sen_terms, by = "id_bioguide") %>%
  mutate(rep_to_sen = last_sen_term > first_rep_term,
         sen_to_rep = first_sen_term < first_rep_term,
         too_late = first_rep_term > '2009-12-31') %>%
  arrange(desc(rep_to_sen), first_sen_term) %>%
  collect()
```
Because of the form of our second analysis, it is easy to make a plot.

```{r}
#| warning: false
pct_rep_then_sen %>%
  mutate(gap = (year(gap) * 12 + month(gap))/12) %>%
  filter(gap <= 15) %>%
  group_by(cohort_century, gap) %>%
  summarize(pct = max(pct),
            .groups = "drop") %>%
  collect() %>%
  mutate(cohort_century = factor(cohort_century)) %>%
  ggplot(aes(x = gap, y = pct, group = cohort_century,
             colour = cohort_century)) +
  geom_line()
```

Note that it would be easy to translate our `dbplyr` code---arguably a more precise solution than the original SQL query---back into SQL (probably using CTEs) if that were desired.

## References {-}
