---
title: "The Gino-Colada Affair"
author: 
  - name: Ian D. Gow
    orcid: 0000-0002-6243-8409
    email: iandgow@gmail.com
abstract: "This short note reproduces analysis found on [Replicability-Index](https://replicationindex.com/2023/09/30/the-gino-colada-affair/)."
date: today
date-format: "D MMMM YYYY"
bibliography: papers.bib
title-block-style: plain
csl: jfe.csl
format: 
    pdf:
        toc: false
        number-sections: true
        colorlinks: true
        mainfont: TeX Gyre Pagella
        sansfont: TeX Gyre Pagella
---

## Study #1

```{r}
#| include: false
library(modelsummary)
library(readxl)
library(dplyr, warn.conflicts = FALSE)
library(broom)
```

```{r}
#| include: false
#| cache: true
t <- tempfile(fileext = ".xlsx")
url <- "https://osf.io/download/2ehzt/"
download.file(url, t)

df <- 
  read_xlsx(t, range = "A1:T102") |>
  rename(id = `P#`) |>
  mutate(Cond = as.factor(Cond),
         disputed = id %in% c(51, 12, 101, 7, 91, 52)) %>%
  mutate(IncomeMinusTax = Income - TaxOnIncome,
         SumDeduction = Deduction1 + Deduction2,
         FinalPayment = IncomeMinusTax + SumDeduction,
         OverReport = `#B` - `#ActuallyCorrect`,
         CheatedOnMatrixTax = OverReport > 0)
```

```{r}
#| include: false
fms <- list()
fms[[1]] <- lm(SumDeduction ~ Cond, data = df)
fms[[2]] <- lm(SumDeduction ~ Cond, data = df, subset = Cond != 0)
fms[[3]] <- lm(SumDeduction ~ Cond, data = df, subset = Cond != 2)
```

```{r}
#| include: false
glance_custom.lm <- function(x, ...) {
    broom::glance(x)
}
```

We can reproduce *the equivalent of* the published results of an ANOVA with the three conditions as categorical predictor variable and deductions as outcome variable *using linear regression*.
Results are reported in column (1) of @tbl-orig.
In addition, the original article reported that each difference between the experimental "signature-on-top" and the two control conditions ("signature-on-bottom", "no signature") was significant.
This is confirmed in columns (2) and (3) of @tbl-orig.

```{r}
#| include: false
fms_no_disp <- list()
fms_no_disp[[1]] <- lm(SumDeduction ~ Cond, data = df, subset = !disputed)
fms_no_disp[[2]] <- lm(SumDeduction ~ Cond, data = df, subset = Cond != 0 & !disputed)
fms_no_disp[[3]] <- lm(SumDeduction ~ Cond, data = df, subset = Cond != 2 & !disputed)
fms_no_disp[[4]] <- lm(SumDeduction ~ Cond == 1, data = df, subset = !disputed)
```

```{r}
#| include: false
f_fixed <- summary(fms_no_disp[[1]])$fstatistic
p_val <- glance(fms_no_disp[[1]])$p.value
```

Next, we can repeat the analysis without rows 67 to 72. 
Results are reported in @tbl-delete.
Without the six contested cases, the results are no longer statistically significant, $F( `r f_fixed[["numdf"]]`, `r f_fixed[["dendf"]]`) = `r prettyNum(f_fixed[["value"]], digits = 3)`$, $p = `r prettyNum(p_val, digits = 3)`$, as seen in column (1) of @tbl-delete.
The comparisons of the experimental group with the two control groups were also statistically significant (see columns (2) and (3) of @tbl-delete).
Combining the two control groups into one and comparing it to the experimental group and presenting the results as a planned contrast would also have produced a significant result (see column (4) of @tbl-delete).

```{r}
#| include: false
avg <- mean(df$SumDeduction)
df_mod <- 
  df |>
  mutate(Cond = case_when(disputed & SumDeduction > avg ~ "1",
                          disputed & SumDeduction < avg ~ "2",
                          .default = Cond))
```

```{r}
#| include: false
fms_fixed <- list()
fms_fixed[[1]] <- lm(SumDeduction ~ Cond, data = df_mod)
fms_fixed[[2]] <- lm(SumDeduction ~ Cond, data = df_mod, subset = Cond != 0)
fms_fixed[[3]] <- lm(SumDeduction ~ Cond, data = df_mod, subset = Cond != 2)
```

```{r}
#| include: false
f_fixed <- summary(fms_fixed[[1]])$fstatistic
p_val <- glance(fms_fixed[[1]])$p.value
```

Of course, the accusation is that she switched rows with low values to the experimental condition and rows with high values to the control condition.
To attempt to reverse this manipulation, we can recode the contested rows 67--69 as signature-at-the-bottom and 70--72 as signature-at-the-top and repeat the analysis.
In this case, there was no evidence that the group means differed from each other, $F( `r f_fixed[["numdf"]]`, `r f_fixed[["dendf"]]`) = `r prettyNum(f_fixed[["value"]], digits = 3)`$, $p = `r prettyNum(p_val, digits = 3)`$.
Results are presented in Column (1) of @tbl-fixed.
Neither comparison of the experimental group with each of the two control groups was statistically significant (see columns (2) and (3) of @tbl-fixed).

```{r}
#| label: tbl-orig
#| echo: false
#| tbl-cap: Reproduction of results from the paper
modelsummary(fms, statistic = "std.error",
             gof_map = c("F", "p.value"))
```

```{r}
#| label: tbl-delete
#| echo: false
#| tbl-cap: Reproduction of results without disputed observations
modelsummary(fms_no_disp, gof_map = c("F", "p.value"))
```

```{r}
#| label: tbl-fixed
#| echo: false
#| tbl-cap: Reproduction of results with corrected data
modelsummary(fms_fixed, statistic = "std.error",
             gof_map = c("F", "p.value"))
```

## Study #2

```{r}
#| include: false
#| cache: true
url <- "https://osf.io/download/ew2ms/"
t <- tempfile()
download.file(url, t)

df_2 <- 
  read_excel(t) |>
  rename(id = `P#`,
         Reported = `#B`,
         Actual = `#ActuallyCorrect`) |>
  mutate(Cheating = Reported - Actual)
```

```{r}
#| include: false
fms_2 <- list()
fms_2[[1]] <- lm(Cheating ~ SignAtTop, data = df_2)
fms_2[[2]] <- lm(SumDeductions ~ SignAtTop, data = df_2)
fms_2[[3]] <- lm(SumEthicsWords ~ SignAtTop, data = df_2)
```

```{r}
#| label: tbl-study-2
#| echo: false
#| tbl-cap: "Study 2: Reproduction of results with OSF data"
modelsummary(dvnames(fms_2), statistic = "statistic",
             gof_map = c("F", "p.value"))
```
The original results from the paper are reported in @tbl-study-2.

From the `calcChain.xml` file, it appears that just three observations (`P#` values `1, 59, 61`) have been moved "out of order" from "sign at the bottom" to "sign at the top".
These observations are (now) in rows 2, 60, and 61 of the OSF spreadsheet.
It seems these changes involved moving a row from the bottom to the top and two rows from the top to the bottom.

```
<c r="I58" i="1"/>
<c r="K58" i="1"/>
<c r="I59" i="1"/>
<c r="K59" i="1"/>
<c r="I2" i="1"/>
<c r="K2" i="1"/>
```
and

```
<c r="I60" i="1"/>
<c r="K60" i="1"/>
<c r="I61" i="1"/>
<c r="K61" i="1"/>
<c r="I3" i="1"/>
<c r="K3" i="1"/>
<c r="I4" i="1"/>
```

Let's see what happens if we move it back?
Results are reported in @tbl-study-2-fixed.

```{r}
#| include: false 
df_2_fixed <-
  df_2 |>
  mutate(SignAtTop = if_else(id %in% c(1, 59, 61), !SignAtTop, SignAtTop))
```


```{r}
#| include: false
fms_2_fixed <- list()
fms_2_fixed[[1]] <- lm(Cheating ~ SignAtTop, data = df_2_fixed)
fms_2_fixed[[2]] <- lm(SumDeductions ~ SignAtTop, data = df_2_fixed)
fms_2_fixed[[3]] <- lm(SumEthicsWords ~ SignAtTop, data = df_2_fixed)
```

```{r}
#| label: tbl-study-2-fixed
#| echo: false
#| tbl-cap: "Study 2: Reproduction of results with corrected data"
modelsummary(dvnames(fms_2_fixed), statistic = "statistic",
             gof_map = c("F", "p.value"))
```

