---
title: "Growth"
author: Ian D. Gow
date: 2025-04-10
date-format: "D MMMM YYYY"
format:
  pdf: 
    colorlinks: true
    geometry:
      - left=2.5cm
      - right=2.5cm
    papersize: a4
    mainfont: TeX Gyre Pagella
    mathfont: TeX Gyre Pagella Math
bibliography: papers.bib
---

```{r}
#| message: false
#| include: false
library(tidyverse)
library(DBI)
library(farr)
library(googlesheets4)
```

```{r}
#| include: false
#| warning: false
# gs4_auth(email = "iandgow@gmail.com")
```

```{r}
#| include: false
start_year <- 2008L
end_year <- 2017L
n_years <- end_year - start_year
```

```{r}
#| include: false
db <- dbConnect(duckdb::duckdb())

funda <- load_parquet(db, schema = "comp", table = "funda")
company <- load_parquet(db, schema = "comp", table = "company")

funda_mod <-
  funda |>
  filter(indfmt == "INDL", datafmt == "STD",
         consol == "C", popsrc == "D")

```

This note was written using [Quarto](https://quarto.org) and compiled with [RStudio](https://posit.co/products/open-source/rstudio/), an integrated development environment (IDE) for working with R.
The source code for this note can be found [here](https://github.com/iangow/notes/blob/main/growth.qmd) and the current version of this PDF can be found [here](https://github.com/iangow/notes/blob/main/growth.pdf)

@fig-growth-ind is a rough replication of Exhibit 9.1 of @Koller:2020aa [p. 157].
While the numbers do not line up exactly, the broad thrust is similar (e.g., *Biotechnology* has high median growth and the widest dispersion).
I think it is helpful to make it possible for readers to see the choices underlying a plot (and to produce their own version with their own choices).

Some observations:

 - While "Compustat" is listed as a data source in the note to Exhibit 9.1 of @Koller:2020aa, it is not clear what sample restrictions were imposed (e.g., size).
 - The industry mapping is not entirely clear. I had ChatGPT [guess](https://chatgpt.com/share/67f82079-ab7c-800b-9448-dcd28f7305f9) a possible mapping. I then cleaned this up a tad and posted it in a [Google Sheets document](https://docs.google.com/spreadsheets/d/1B0YvlXC1Mw7fM8IMLe_DhYh1n1E1Et-3__gi9vZoIwY/edit?usp=sharing).
 I would proably need to go through carefully "by hand" to check that no firms are inadvertedly omitted or put in the wrong industry group.
 - I like the way the word "Industry" appears (in **bold**) at the top of the y-axis labels (industries) in Exhibit 9.1 of @Koller:2020aa.
 I couldn't quite replicate that here.
 - I couldn't get the legend seen in Exhibit 9.1 of @Koller:2020aa to look good in @fig-growth-ind.
 - I had ChatGPT [help me](https://chatgpt.com/share/67f82079-ab7c-800b-9448-dcd28f7305f9) with the creation of the markers at each quantile.
 - Updating @fig-growth-ind to cover different periods is quite easy.
 - Other plots in Chapter 9 of @Koller:2020aa might be amenable to creation in the same way.

```{r}
#| echo: false
#| warning: false
gs_inds <- 
  read_sheet("1B0YvlXC1Mw7fM8IMLe_DhYh1n1E1Et-3__gi9vZoIwY") |>
  separate_longer_delim(c(codes, field), ",") |> 
  mutate(codes = str_trim(codes)) |>
  copy_to(db, df = _, name = "gs_inds", overwrite = TRUE) |>
  pivot_wider(values_from = "codes", names_from = "field") |>
  compute()
```
```{r}
#| echo: false
industries_mck <-
  company |>
  filter(fic == "USA") |>
  left_join(
    gs_inds |> 
      select(industry_mck, gsubind), by = "gsubind") |>
  rename(industry_mck_subind = industry_mck) |>
  left_join(
     gs_inds |> 
      select(industry_mck, gind), by = "gind") |>
  rename(industry_mck_ind = industry_mck) |>
  left_join(
    gs_inds |> 
      select(industry_mck, ggroup), by = "ggroup") |>
  rename(industry_mck_group = industry_mck) |>
  mutate(industry_mck = coalesce(industry_mck_subind, 
                                 industry_mck_ind, 
                                 industry_mck_group)) |>
  select(gvkey, industry_mck, ggroup, gind, gsubind) |>
  compute()
```

```{r}
#| echo: false
years <-
  tibble(period = c("start_year", "end_year"), 
         fyear = c(start_year, end_year)) |>
  copy_to(db, df = _, name = "years", overwrite = TRUE)


growth_data <-
  funda_mod |> 
  inner_join(years, by = join_by(fyear)) |>
  select(gvkey, period, sale) |> 
  pivot_wider(names_from = "period", values_from = "sale") |> 
  mutate(cagr = 100 * ((end_year / start_year) ^ (1 / n_years) - 1)) |>
  inner_join(industries_mck, by = join_by(gvkey)) |>
  compute()

growth_stats <- 
  growth_data |>
  group_by(industry_mck) |>
  summarise(
    q25 = approx_quantile(cagr, 0.25),
    median = approx_quantile(cagr, 0.50),
    q75 = approx_quantile(cagr, 0.75)
  ) |>
  collect()
```

```{r}
#| echo: false
#| label: fig-growth-ind
#| fig-cap: Variation in revenue growth by industry
#| fig-cap-location: top
growth_stats |>
  filter(!is.na(industry_mck)) |>
  arrange(median) |>
  mutate(industry_mck = factor(industry_mck, levels = industry_mck)) |>
  ggplot(aes(x = median, y = industry_mck)) +
  geom_pointrange(aes(xmin = q25, xmax = q75), color = "steelblue", 
                  shape = 22, size = .5, fill = "white") +
  # Square at Q1
  geom_point(aes(x = q25), shape = 15, size = 2.5, color = "steelblue") +
  # Circle at Q3
  geom_point(aes(x = q75), shape = 16, size = 2.5, color = "steelblue") +
  theme_minimal() +
  labs(
    # title = "Annual revenue growth rate, %",
    x = "2008-2017 average, %",
    y = "Industry"
  ) +
  labs(y = NULL) +
  theme(
    axis.text.y = element_text(
      size = 8,
      hjust = 0)
  )
```
# References {-}
