---
title: "Macro homework"
author: 
  - name: Ian D. Gow
    orcid: 0000-0002-6243-8409
    email: iandgow@gmail.com
date: 2025-02-01
number-sections: true
format:
  html:
    default
  pdf: 
    include-in-header:
      text: |
        \usepackage[group-digits = integer, group-separator={,}, group-minimum-digits = 4]{siunitx}
        \usepackage{scrextend}
        \deffootnote{1.6em}{1.6em}{\thefootnotemark.\enskip}
        \addtokomafont{disposition}{\rmfamily}
    colorlinks: true
    geometry:
      - left=2cm
      - right=2cm
    papersize: a4
    mainfont: TeX Gyre Pagella
    mathfont: TeX Gyre Pagella Math
bibliography: papers.bib
---

```{r}
#| include: false
library(tidyverse)
```

```{r}
#| cache: true
#| include: false
get_series_names <- function(url) {
  df_names <- read_csv(url, n_max = 1,
                       show_col_types = FALSE)
  t(df_names[1, ]) |> 
    as_tibble(rownames = "desc") |> 
    rename(series = `V1`)
}

get_data <- function() {
  url <- str_c("https://www.ons.gov.uk/file?uri=/",
               "employmentandlabourmarket/peopleinwork/labourproductivity/",
               "datasets/labourproductivity/current/prdy.csv")
  df <- read_csv(url, skip = 6, show_col_types = FALSE)

  series_details <- get_series_names(url)
  
  colnames(df) <- series_details$series
  return(df)
}

df <- get_data()
```

```{r}
#| include: false
qtr_to_date <- function(yearq) {
  year <- str_extract(yearq, "^([\\d]+)", 1)
  quarter <- as.integer(str_extract(yearq, "Q(\\d)", 1))
  month <- quarter * 3
  date <- as.Date(str_c(year, "-", month, "-01"))
  date + months(1) - days(1)
}
```

```{r}
#| include: false
labour_data <-
  df |> 
  select(CDID, LZVB, A4YM) |>
  filter(!is.na(LZVB), str_detect(CDID, "[\\d]{4} Q[\\d]")) |>
  pivot_longer(-CDID, names_to = "series", values_to = "value") |>
  mutate(date = qtr_to_date(CDID))
```

# Data and the current economy

## 1. Labour productivity
Download data on UK "Labour Productivity" from weblearn or from the [website of the Office for
National Statistics](http://www.ons.gov.uk/). 
The correct data should have the name PRDY.

  i. Using a spreadsheet or other computer program, make a graph of the column labelled
"UK Whole Economy: Output per hour worked SA: Index 2018 = 100" over time (code =
LZVB). 
Make sure you only use the quarterly data from 1971Q1 to as close to the present
day as possible.

See @fig-lzvb.

```{r}
#| label: fig-lzvb
#| fig-cap: Output per hour (LZVB) over time
labour_data |>
  filter(series == "LZVB") |>
  ggplot(aes(x = date, y = value, colour = series)) +
  geom_line()
```

  ii. Discuss the shape of the graph after the 2008 financial crisis.
  
After 2008, it apppears from @fig-lzvb that producitivity growth (ie, increases in output per hour worked) levels off (ie, its growth rate decreases).
  
  iii. Add a trend line to the graph to more precisely pin down how UK labour productivity
behaved after the 2008 financial crisis. 
  What is the big picture? 
Why does the trend line in your graph look different to that produced [by the ONS](http://visual.ons.gov.uk/productivity-puzzle)? 
Which graph gives the most accurate description of reality, yours or the one by the ONS?

Not clear which trendline the question is asking for, but I add two trendlines: one for pre-crisis and another for post-crisis.^[I label anything in 2008 or later as post-crisis.]
The graph produced by the ONS effectively projects the pre-crisis trend into the post-crisis period.
It is not entirely clear what "reality" is intended to be described by the ONS plot.
In some sense, the "reality" intended to be described by the ONS plot might be the **counterfactual** world in which the pre-crisis trend continued into the post-crisis period.
But clearly @fig-lzvb-trend better describes the reality of the world actually observed post-crisis.

```{r}
#| label: fig-lzvb-trend
#| fig-cap: Output per hour (LZVB) with trendlines
df_lzvb <- 
  labour_data |>
  filter(series == "LZVB") |>
  mutate(post_crisis = date >= "2008-01-01") 

df_lzvb |>
  ggplot(aes(x = date, y = value, colour = post_crisis)) +
  geom_line() +
  geom_smooth(data = df_lzvb |> filter(post_crisis), 
              method = "lm",
              formula = 'y ~ x',
              se = FALSE) +
  geom_smooth(data = df_lzvb |> filter(!post_crisis), 
              method = "lm",
              formula = 'y ~ x',
              se = FALSE)
```

   iv. Why might UK labour productivity have behaved as it has since the 2008 financial crisis?
Have a look at the [Bank of England's Quarterly Bulletin 2014 Q2](https://www.bankofengland.co.uk/-/media/boe/files/quarterly-bulletin/2014/the-uk-
productivity-puzzle.pdf) and the [2014 speech by
Martin Weale "The UK productivity puzzle: an international perspective"](https://www.bankofengland.co.uk/-/media/boe/files/news/2014/december/the-uk-
productivity-puzzle-an-international-perspective-speech-by-martin-weale.pdf) for background
reading.
  v. Examine also the series "Output per Worker: Whole Economy SA: Index 2018=100: UK"
(code = A4YM). Are you surprised by the behaviour of the productivity series in 2020 Q2
when the pandemic hit? If so, why? If not, why not?

In @fig-both, I also include the "Output per Worker" series.
There was a much-sharper decrease during Covid for this additional series than there was for output per hour worked, consistent with a sharp decrease in hours worked duing the early part of the Covid pandemic.

```{r}
#| label: fig-both
#| fig-cap: Output per hour (LZVB) and per worker (A4YM) over time
labour_data |>
  ggplot(aes(x = date, y = value, colour = series)) +
  geom_line()
```

## 2. Real versus nominal variables

If there is one place in the world that does not understand inflation it is Hollywood.
Even after being open for less than a month, *Star Wars Ep. VIII: The Last Jedi* had moved to 8th on the all-time, highest-grossing domestic movie list.^[<http://variety.com/2017/film/news/box-office-star-wars-last-jedi-jumanji-2-1202650420/>] 
And the 2015 release of *Star Wars Ep. VII: The Force Awakens* shattered every box office record.^[<https://www.cinemablend.com/new/Star-Wars-Force-Awakens-Poised-Shatter-Every-Box-Office-Record-We-Have-70996.html>]

How do reported worldwide box office takings of $1,370,011,043 for *Star Wars Ep. VII: The Force
Awakens* in 2015 really compare with the $786,535,665 brought in by the original *Star Wars Ep.
IV: A New Hope* in 1977?
Download data for CPI inflation from <https://research.stlouisfed.org/fred2/> to calculate the box
office takings of Ep. IV in 2015 dollars and the box office takings of Ep. VII in 1977 dollars.
Which film had the greater takings?
(Bonus for Star Wars fans: which do you think was the better film?)

# Models in economics

Suppose a representative agent has a utility function involving consumption and labour supply
of the form $U = \log(C) - 2L^2$ where $\log (C)$ is the natural logarithm.
 
 i) The agent is a 'yeoman farmer', that produces their own output with a production
function $𝐶 = Y = AL^{\alpha}$, where $Y$ is output per head.
Derive the optimal level of labour supply, and comment on what it implies for the relationship between labour supply and productivity growth.

 ii) The agent is instead a worker, who receives a real wage $w$.
Their budget constraint is $C = wL + \pi$, where $\pi$ are profits distributed by firms, and both $w$ and $\pi$ are assumed to be exogenous by the worker.
Derive an expression for optimal labour supply as a function of $w$ and $C$.
 
 iii) An individual, perfectly competitive firm maximises profits $\pi = A N^n - w N$, where $N$ is the number of workers it employs.
Derive the labour demand curve for this firm.
What is the relationship between the marginal product of labour and the real wage?

 iv) Since each worker works $L$ hours and each firm wishes to employ a total of $N$ hours, equilibrium requires that $L = nN$ where $n$ is the number of firms per worker (or, equivalently, the reciprocal of the number of workers per firm).
As consumption per head is equal to output per head, $C = nY = n AN^n$.
Use this, plus the expression for the real wage, in the labour supply equation to derive the equilibrium value for $L$ in terms of model parameters.
Compare this to your answer to (i), and comment.

 v) Suppose the number of agents in the economy increases (because of immigration, for example), but the new agents are just like the existing ones.
As a result, the number of firms $n$ falls.
Show what happens to real wages, output per head and consumption per head.

 vi) By deriving an expression for profits per firm in terms of $L$ and $n$, comment on what might happen in the long run as $n$ changes.
