---
title: "A quick look at N-CSR and N-CEN filings"
author: Ian D. Gow
date: 2024-09-19
date-format: "D MMMM YYYY"
format:
  html:
    colorlinks: true
  pdf: 
    colorlinks: true
    geometry:
      - left=2.5cm
      - right=2.5cm
    papersize: a4
    mainfont: TeX Gyre Pagella
    mathfont: TeX Gyre Pagella Math
bibliography: papers.bib
---


In writing this note, I use the packages listed below.^[Execute `install.packages(c("tidyverse", "DBI", "duckdb", "farr"))` within R to install all the packages you need to run the code in this note.]
This note was written using [Quarto](https://quarto.org) and compiled with [RStudio](https://posit.co/products/open-source/rstudio/), an integrated development environment (IDE) for working with R.
The source code for this note is available [here](https://raw.githubusercontent.com/iangow/notes/main/ncsr.qmd).

```{r}
#| include: false
options(width = 75,
        tibble.width = 75,
        pillar.print_min = 5)
```

```{r}
#| message: false
library(tidyverse)
library(DBI)
library(farr)
```
I have SEC filing index files downloaded as discussed [here](https://iangow.github.io/far_book/glms.html#appendix-maintaining-a-repository-of-sec-index-files).

```{r}
#| eval: true
db <- dbConnect(duckdb::duckdb())

sec_index <- load_parquet(db, "sec_index*", "edgar")
```

It's often helpful to look for variants of a `form_type`.

```{r}
sec_index |>
  filter(str_detect(form_type, "CSR")) |>
  count(form_type, sort = TRUE) |>
  collect()
```

Here I just grab a sample of N-CSR filings.

```{r}
sample_filings <-
  sec_index |> 
  filter(form_type == "N-CSR") |>
  arrange(desc(date_filed)) |>
  collect(n = 10)
```

I make `browse_sec()`, a small function for browsing filings on the SEC website.

```{r}
browse_sec <- function(file_name) {
  url <- str_replace(file_name, 
                     "([0-9]{10})-([0-9]{2})-([0-9]{6})\\.txt$",
                     "\\1\\2\\3/\\1-\\2-\\3-index.htm")
  url <- str_c("https://www.sec.gov/Archives/", url)
  browseURL(url)
}
```

And I use `browse_sec()` to look at the first filing.

```{r}
browse_sec(sample_filings$file_name[1])
```

I couldn't see any director names there, but I did find them in a "N-CEN" filing around the same time.

It's often helpful to look for variants of a `form_type`.

```{r}
sec_index |>
  filter(str_detect(form_type, "CEN")) |>
  count(form_type, sort = TRUE) |>
  collect()
```
This seems much rarer than N-CSR filings. Perhaps this a new form type.

Anyway, I grab a sample of these ...

```{r}
sample_cen_filings <-
  sec_index |> 
  filter(form_type == "N-CEN") |>
  arrange(desc(date_filed)) |>
  collect(n = 10)
```

... and look at the first one.

```{r}
browse_sec(sample_cen_filings$file_name[1])
```

This filing has the directors' names in XML, so pretty easy to extract.
