---
title: "Retail sales"
author: 
  - name: Ian D. Gow
    orcid: 0000-0002-6243-8409
    email: iandgow@gmail.com
date: 2025-02-03
number-sections: true
format:
  html:
    default
  pdf: 
    include-in-header:
      text: |
        \usepackage[group-digits = integer, group-separator={,}, group-minimum-digits = 4]{siunitx}
        \usepackage{scrextend}
        \deffootnote{1.6em}{1.6em}{\thefootnotemark.\enskip}
        \addtokomafont{disposition}{\rmfamily}
        \sisetup{output-decimal-marker = {,}}
    colorlinks: true
    geometry:
      - left=2cm
      - right=2cm
    papersize: a4
    mainfont: TeX Gyre Pagella
    mathfont: TeX Gyre Pagella Math
bibliography: papers.bib
---

:::{#tip-pkgs .callout-tip text-align="left"}
The code in this chapter uses the packages listed below.
For instructions on how to set up your computer to use the code found in this book, see @sec-install.
Quarto templates for the exercises below are available on [GitHub](https://github.com/iangow/far_templates/blob/main/README.md).
:::

```{r}
#| warning: false
library(tidyverse)
library(modelsummary)
library(dbplyr)
library(DBI)
```

```{r}
#| eval: false
#| include: false
library(readabs)

Sys.setenv(R_READABS_PATH = "~/Downloads/")

retail_industries <- tribble(
  ~industry, ~ind_level,
  "Supermarket and grocery stores", 2,
  "Liquor retailing", 2,
  "Other specialised food retailing", 2,
  "Food retailing", 1,
  "Furniture, floor coverings, houseware and textile goods retailing", 2,
  "Electrical and electronic goods retailing", 2,
  "Hardware, building and garden supplies retailing", 2,
  "Household goods retailing", 1,
  "Clothing retailing", 2,
  "Footwear and other personal accessory retailing", 2,
  "Clothing, footwear and personal accessory retailing", 1,
  "Department stores", 1,
  "Newspaper and book retailing", 2,
  "Other recreational goods retailing", 2,
  "Pharmaceutical, cosmetic and toiletry goods retailing", 2,
  "Other retailing n.e.c.", 2,
  "Other retailing", 1,
  "Cafes, restaurants and catering services", 2,
  "Takeaway food services", 2,
  "Cafes, restaurants and takeaway food services", 1,
  "Total (Industry)", 0) |>
  mutate(parent_industry = if_else(ind_level == 1, industry, NA)) |>
  fill(parent_industry, .direction = "up") |>
  mutate(parent_industry = if_else(ind_level == 1, "Total (Industry)", 
                                   parent_industry)) 

table <- "TABLE 11. Retail Turnover, State by Industry Subgroup, Original"

abs_data <-
  read_abs("8501.0") |>
  filter(table_title == table) |>
  separate_wider_regex(series,
                       c(label = "^.*", ";\\s+", 
                         state = ".*?", "\\s*;\\s*",
                         industry = ".*?", "\\s*;$")) |>
  mutate(label = str_trim(label)) |>
  select(industry, state, date, value) |>
  rename(sales = value) |>
  arrange(date, industry, state)

retail_sales <-
  abs_data |>
  left_join(retail_industries, join_by(industry)) |>
  write_csv("data/retail_sales.csv") 
```

The data we focus on in this chapter come from the [Australian Bureau of Statistics](https://www.abs.gov.au) (ABS).
The ABS describes itself as "Australia's national statistical agency and an official source of independent, reliable information."
We discuss the source of the data in more detail in @tip-abs.

The data are stored as a **comma-separated value** (CSV) file format in `data/retail_sales.csv` in the repository you downloaded for this course.
There is a good chance that, if Microsoft Excel is installed on your computer, double-clicking on this file on your computer will open it in Excel.
To understand the contents of the file, it's actually better to open it in a text editor, such as Notepad (Windows) or TextEdit (MacOS).
CSV files are a common format for sharing data, as it is a simple format that many software packages can handle.
[Chapter 7](https://r4ds.hadley.nz/data-import.html) of *R for Data Science* discusses CSVs in more detail.

Here we use `read_csv()` from the Tidyverse to load the data into R.

```{r}
retail_sales <- read_csv("data/retail_sales.csv")
```

As can be seen in the output above, `read_csv()` will attempt to infer a number of features of the data.
For example, the first row of the CSV file (`industry,state,date,sales,ind_level,parent_industry`) is assumed to provide the names the columns of the resulting data frame.
Additionally, `read_csv()` will guess the appropriate type for each column.
From the output above, we can see that `industry`, `state`, and `parent_industry` are imported as character columns, that `date` has type date, and that `sales` and `ind_level` have type `dbl` ("double" being the R type for storing real numbers).

The data in `retail_sales` are monthly estimates of turnover [sales in dollars] and volumes for retail businesses, including store and online sales, by industry group and industry subgroup.
According to [the ABS](https://www.abs.gov.au/methodologies/retail-trade-australia-methodology/dec-2024), "estimates of turnover are compiled from the monthly Retail Business Survey.
About 700 'large' businesses are included in the survey every month, while a sample of about 2,700 'smaller' businesses is selected."

# Understanding the industries

According to [the ABS,](https://www.abs.gov.au/methodologies/retail-trade-australia-methodology/jul-2024#defining-retail-trade) the statistics from which `retail_sales` are derived are presented at two levels of detail: *industry group*, "the broadest industry level comprising 6 industry groups", and *industry subgroup*, which is "the most detailed industry level comprising 15 industry subgroups."
In `retail_sales`, industry groups are associated with `ind_level` equal to 1 and industry subgroups are associated with `ind_level` equal to 2.
For each, industry subgroup, the industry group to which it belongs is identified in `parent_industry`.

We can take a peek at the first few rows of data by typing `retail_sales` in the R console:

```{r}
retail_sales
```


We can also get a succinct take on the data using `summary()`:

```{r}
summary(retail_sales)
```
We start by investigating the data in `industry`, `ind_level`, and `parent_industry`.
Below we see that `Total (Industry)` is the parent industry for all industries having `ind_level` equal to one and that `parent_industry` is missing whenever `ind_level` is zero.
Note that only five distinct industries are listed under `parent_industry` when `ind_level` is two.

```{r}
retail_sales |> 
  count(parent_industry, ind_level) |>
  arrange(ind_level)
```

We can identify the missing industry group by using an **anti-join** (`anti_join()`) and `join_by(industry == parent_industry)`.
The output of following query identifies the problem case as "Department stores".

```{r}
retail_sales |>
  filter(ind_level == 1) |>
  anti_join(retail_sales, by = join_by(industry == parent_industry)) |>
  count(industry)
```

Because the function `anti_join(x, y)` returns all rows in `x` that don’t have a match in `y`, it is useful for finding **implicit missing values**.^[TODO: Discuss *implicit missing values* in an earlier section on `left_join()`.]

Additionally, while the ABS website describes the data as "comprising 15 industry subgroups", we see only 14 subgroups in the data:

```{r}
retail_sales |> 
  distinct(industry, ind_level) |> 
  count(ind_level)
```

What explains these discrepancies?
The natural explanation is that *Department stores* is an industry group that has only one subgroup, itself.
In effect, *Department stores* could be considered as having `ind_level` equal to both one and two.

A natural question is whether sales of the subgroups add up to the sales of the industry groups.^[Once we recognize that *Department stores* is its own parent industry, it is trivial that its own sales will "add up" in the sense here, so we don't need to consider that industry for present purposes.]
To answer this question, I first collect data---the sum of sales by state and year---on industry groups directly and store them in `level_1_df`.

```{r}
level_1_df <-
  retail_sales |> 
  mutate(year = year(date)) |>
  filter(ind_level == 1, year < 2024) |>
  group_by(state, year, industry, ind_level) |>
  summarize(no_missing = as.integer(all(!is.na(sales))),
            sales = sum(sales, na.rm = TRUE),
            .groups = "drop")
```

I then collect the same data on industry groups directly by aggregating the applicable industry subgroups and store them in `level_2_df`.

```{r}
level_2_df <-
  retail_sales |> 
  mutate(year = year(date)) |>
  filter(ind_level == 2, year < 2024) |>
  group_by(state, year, parent_industry, ind_level) |>
  summarize(no_missing = as.integer(all(!is.na(sales))),
            sales = sum(sales, na.rm = TRUE),
            .groups = "drop") |>
  rename(industry = parent_industry)
```

Finally, I combine the data from both of these data frames into one data frame using `union_all()` and save the result in `combined_levels`.

```{r}
combined_levels <-
  level_1_df |>
  union_all(level_2_df) |>
  mutate(year = as.character(year),
         ind_level = as.character(ind_level))
```

I will use `datasummary()` from the `modelsummary` package to display summary statistics.
I create `Sum()` to sum up values while removing `NA` values and `pretty_num()` to format the numbers in the way that I want.

```{r}
Sum <- function(x) sum(as.double(x), na.rm = TRUE)
pretty_num <- function(x) prettyNum(as.integer(x), format = "f", 
                                    big.mark = ",")
```

I create alternative shorter names for two industries that have names that are too long to fit in the tables I want to display.

```{r}
short_names <-
  tribble(
  ~industry, ~short_ind,
  "Cafes, restaurants and takeaway food services", "Cafes, restaurants, etc",
  "Clothing, footwear and personal accessory retailing", 
    "Clothing, footwear, accessories")
```

In @tbl-reconc, I display the total sales for each industry group calculated in two ways: once by summing up the values for the industry group and again by summing up the values for the subgroups comprised by the group.

```{r}
#| label: tbl-reconc
#| tbl-cap: "Reconciliation of sales by group and sub-groups: Total (State)"
combined_levels |>
  left_join(short_names, by = "industry") |>
  mutate(industry = coalesce(short_ind, industry)) |>
  filter(year >= 2019, 
         state == "Total (State)",
         industry != "Department stores") |>
  datasummary(industry * ind_level ~ sales * Sum * year,
              data = _, fmt = pretty_num,
              digits = 1)
```

With any data set it is important to understand the meaning of any explicit missing values.
For the five industry groups (excluding *Department stores* for the reasons discussed above), we have complete information at level 1 for a year if there are 12 months of data for each industry group at level 1.
And we have complete information at level 2 for a year if there are 12 months of data for each sub-group at level 2.
From @tbl-non-missing, we see that larger states have complete data for all five industries at both levels one and two, but that smaller states such as Tasmania and Northern Territory have missing values at both levels.^[Of course, Northern Territory is not a *state*, but a territory.]

```{r}
#| label: tbl-non-missing
#| tbl-cap: "Non-missing values by state and year and industry level"
combined_levels |>
  filter(year >= 2019, 
         industry != "Department stores") |>
  datasummary(state * ind_level ~ no_missing * Sum * year,
              data = _, fmt = pretty_num)
```

## Exercises

1. Describe in words what the query above using `anti_join()` is doing.

2. Suppose we wanted to focus our analysis on industry subgroups, but wanted to retain "Department stores" as part of this.
Write code to create `ind_sub_groups`, a data set comprising data from `retail_sales` that we could use for this purpose.

```{r}
#| label: tbl-reconc-level-2
#| tbl-cap: "Sales by industry subgroup"
#| include: false
combined_levels |>
  mutate(ind_level = if_else(industry == "Department stores", "2", ind_level)) |>
  filter(ind_level == "2",
         year >= 2019) |>
  datasummary(industry * ind_level ~ sales * Sum * year,
              data = _, fmt = pretty_num)
```

3. What industry group has missing values at level 2 for Queensland, as seen in @tbl-non-missing?
Which subgroups are affected?

```{r}
#| include: false
combined_levels |>
  filter(year >= 2019, 
         state == "Queensland",
         industry != "Department stores") |>
  datasummary(industry * ind_level ~ sales * Sum * year,
              data = _, fmt = pretty_num)
```

4. Produce a version of @tbl-reconc but focused on Queensland instead of all of Australia (`Total (State)`).
Do you see an issue related to the issue flagged in the previous question?

```{r}
#| label: tbl-reconc-qld
#| tbl-cap: "Reconciliation of sales by group and sub-groups: Queensland"
#| include: false
combined_levels |>
  left_join(short_names, by = "industry") |>
  mutate(industry = coalesce(short_ind, industry)) |>
  filter(year >= 2019, 
         state == "Queensland",
         industry != "Department stores") |>
  datasummary(industry * ind_level ~ sales * Sum * year,
              data = _, fmt = pretty_num,
              digits = 1)
```

5. Produce a version of @tbl-reconc but focused on Tasmania instead of all of Australia (`Total (State)`).
Do you see an issue related to the issue flagged in the previous question?

```{r}
#| label: tbl-reconc-tas
#| tbl-cap: "Reconciliation of sales by group and sub-groups: Tasmania"
#| include: false
combined_levels |>
  left_join(short_names, by = "industry") |>
  mutate(industry = coalesce(short_ind, industry)) |>
  filter(year >= 2019, 
         state == "Tasmania",
         industry != "Department stores") |>
  datasummary(industry * ind_level ~ sales * Sum * year,
              data = _, fmt = pretty_num,
              digits = 1)
```

# Time-series analysis

What are time-series data?

## Plotting time-series data

```{r}
start_date <- as.Date("2010-01-01")
end_date <- as.Date("2024-11-30")

df_group_plot <-
  retail_sales |>
  left_join(short_names, by = "industry") |>
  mutate(industry = coalesce(short_ind, industry)) |>
  filter(state == "Total (State)", ind_level == 1,
         between(date, start_date, end_date))
```

Figures [-@fig-all-ind] and [-@fig-all-ind2] plot sales from `r format(start_date, "%B %Y")` until `r format(end_date, "%B %Y")` by industry group.

```{r}
#| label: fig-all-ind
#| fig-cap: Retail sales by industry using `facet_wrap()`
df_group_plot |>
  ggplot(aes(x = date, y = sales / 1e3)) +
  ylab("Sales ($ billion)") +
  xlab("Month") +
  geom_line() +
  facet_wrap(industry ~ ., ncol = 2, scales = "free_y")
```

```{r}
#| label: fig-all-ind2
#| fig-cap: Retail sales by industry
df_group_plot |>
  ggplot(aes(x = date, y = sales / 1e3, colour = industry)) +
  ylab("Sales ($ billion)") +
  xlab("Month") +
  geom_line() +
  theme(legend.position = "bottom")
```
### Seasonality

```{r}
#| label: fig-season
#| fig-cap: Seasonality
#| fig-height: 5
#| fig-width: 6
retail_sales |>
  filter(industry == "Cafes, restaurants and takeaway food services",
         state == "Total (State)",
         between(date, start_date, end_date)) |>
  mutate(month = month(date, label = TRUE),
         year = year(date)) |>
  ggplot(aes(x = month, y = sales / 1e3, fill = month)) +
  ylab("Sales ($ billion)") +
  xlab("Month") +
  geom_col() +
  facet_wrap(. ~ year, ncol = 3) +
  theme(legend.position = "none") +
  scale_x_discrete(labels = \(x) str_sub(x, 1, 1))
```

#### Exercises

:::{#tip-backends .callout-tip text-align="left"}

# Alternative backends

So far, we have done many calculations using `mutate()` and `summarize()`, combined data frames using functions like `inner_join()`, or focused on certain rows or columns using `filter()` or `select()`.
We have generally applied the functions to data in data frames, which we understand to comprise vectors of data stored in the memory (RAM) of our computer.
However, as you work more with data, you are likely to encounters situations where this approach needs to be modified.

For example, the data you want to work with might be stored on a remote computer, such as a database server or a Spark cluster.
Alternatively, you might have the data stored on disk, but it either would take too long to load or simply cannot fit into RAM.

It turns out that there are packages that can help us work with this kind data with little change to how we have worked with data so far.


Backend      | Native language  | R package    | `dplyr` interface |
-------------|------------------|--------------|-------------------|
SQL database | SQL              | `DBI`        | `dbplyr`          |
Arrow        | Apache Arrow API | `arrow`      | `arrow`           |
Spark        | Spark SQL, etc.  | `splarklyr`  | `splarklyr`       |
`data.table` | `data.table`     | `data.table` | `dtplyr`          |

In this book, in addition to using `dplyr` with data frames, we explore the use of SQL database as an alternative backend.

Advantages

 - Moving processing to where the data are located
 - Moving processing to a more powerful computer
 - Avoiding loading of data into RAM
 - Using a more powerful backend
 - Using additional functions

:::

## Time windows

Many analyses in business and economics involve calculations over **time windows**.
Some examples:

 - Returns on stocks after their initial public offerings (IPOs)
 - Comparing sales or profits for the year to date with previous years
 - Smoothing sales over time using moving averages

```{r}
db <- dbConnect(duckdb::duckdb())
retail_sales_db <-
  retail_sales |>
  copy_to(db, df = _, overwrite = TRUE)
```

### Year-to-date (YTD)

```{r}
df_group_ytd_plot <-
  retail_sales_db |> 
  filter(str_detect(parent_industry, "^Cafes")) |>
  mutate(year = year(date)) |>
  group_by(industry, state, year) |>
  window_order(date) |>
  mutate(sales_ytd = cumsum(sales)) |>
  filter(year >= 2020) |>
  collect() |>
  mutate(year = as.character(year),
         month = month(date, label = TRUE))
```

```{r}
#| label: fig-all-ind-ytd
#| fig-cap: Year-to-date retail sales by industry
df_group_ytd_plot |>
  filter(state %in% c("Victoria", "Western Australia")) |>
  ggplot(aes(x = month, y = sales_ytd / 1e3, color = year, group = year)) +
  ylab("Sales, year-to-date ($ billion)") +
  xlab("Month") +
  geom_line() +
  facet_wrap(industry ~ state, ncol = 2, scales = "free_y") +
  theme(legend.position = "bottom")
```


### Moving averages

```{r}
df_group_ma_plot <-
  retail_sales_db |>
  left_join(short_names, by = "industry", copy = TRUE) |>
  mutate(industry = coalesce(short_ind, industry)) |>
  group_by(industry, state) |>
  window_order(date) |>
  window_frame(-11, 0) |>
  mutate(sales_ma = cummean(sales),
         count_ma = cumsum(as.integer(!is.na(sales)))) |>
  filter(state == "Total (State)", ind_level == 1,
         between(date, start_date, end_date),
         count_ma >= 12) 
```

```{r}
#| label: fig-all-ind-ma
#| fig-cap: Retail sales (12-month moving average) by industry
df_group_ma_plot |>
  ggplot(aes(x = date, y = sales_ma / 1e3)) +
  ylab("Sales, 12-month moving average ($ billion)") +
  xlab("Month") +
  geom_line() +
  facet_wrap(industry ~ ., ncol = 2, scales = "free_y")
```

### Economic shocks

```{r}
liquor <-
  retail_sales |>
  filter(state %in% c("Victoria", "Western Australia"),
         industry == "Liquor retailing") |>
  filter(between(date, 
                 as.Date("2018-01-01"), 
                 as.Date("2024-08-31"))) |>
  mutate(year = year(date),
         month = month(date, label = TRUE))
```

```{r}
#| label: fig-vic-wa-liq
#| fig-cap: Liquor retailing sales (normalised by 2018 average monthly sales)
#| fig-height: 8
liquor |>
  filter(year == 2018) |>
  group_by(state) |>
  summarize(scale_factor = mean(sales)) |>
  inner_join(liquor, by = "state") |>
  mutate(sales_normalised = sales / scale_factor) |>
  ggplot(aes(x = month, y = sales_normalised, fill = state)) +
  geom_col(position = "dodge") +
  facet_wrap(year ~ ., ncol = 1) +
  theme(legend.position = "bottom") +
  theme(plot.title.position = "plot")
```

:::{#tip-abs .callout-tip text-align="left"}

### Where to find the data

The data used to create `retail_sales` come from the ABS

```{r}
Sys.setenv(R_READABS_PATH = "~/Downloads/")
library(readabs)
```

```{r}
#| cache: true
table <- "TABLE 11. Retail Turnover, State by Industry Subgroup, Original"

abs_data <-
  read_abs("8501.0") |>
  filter(table_title == table) |>
  separate_wider_regex(series,
                       c(label = "^.*", ";\\s+", 
                         state = ".*?", "\\s*;\\s*",
                         industry = ".*?", "\\s*;$")) |>
  mutate(label = str_trim(label)) |>
  select(industry, state, date, value) |>
  rename(sales = value) |>
  arrange(date, industry, state)

abs_data
```

One thing that can be observed in the output above is that there is no indication of which industries are groups and which are subgroups.
Also there is no indication of how the subgroups map to groups.
To handle this, I created the following data frame to map `industry` to `ind_level`.

```{r}
retail_industries <- tribble(
  ~industry, ~ind_level,
  "Supermarket and grocery stores", 2,
  "Liquor retailing", 2,
  "Other specialised food retailing", 2,
  "Food retailing", 1,
  "Furniture, floor coverings, houseware and textile goods retailing", 2,
  "Electrical and electronic goods retailing", 2,
  "Hardware, building and garden supplies retailing", 2,
  "Household goods retailing", 1,
  "Clothing retailing", 2,
  "Footwear and other personal accessory retailing", 2,
  "Clothing, footwear and personal accessory retailing", 1,
  "Department stores", 1,
  "Newspaper and book retailing", 2,
  "Other recreational goods retailing", 2,
  "Pharmaceutical, cosmetic and toiletry goods retailing", 2,
  "Other retailing n.e.c.", 2,
  "Other retailing", 1,
  "Cafes, restaurants and catering services", 2,
  "Takeaway food services", 2,
  "Cafes, restaurants and takeaway food services", 1,
  "Total (Industry)", 0) |>
  mutate(parent_industry = if_else(ind_level == 1, industry, NA)) |>
  fill(parent_industry, .direction = "up") |>
  mutate(parent_industry = if_else(ind_level == 1, "Total (Industry)", 
                                   parent_industry)) 
```

```{r}
retail_sales <-
  abs_data |>
  left_join(retail_industries, join_by(industry)) |>
  write_csv("data/retail_sales.csv") 
```
:::
