---
title: "ASX End of Day Prices"
author: Ian D. Gow (based on SIRCA document)
date: 2024-05-23
date-format: "D MMMM YYYY"
format:
  html:
    colorlinks: true
  pdf: 
    colorlinks: true
    geometry:
      - left=2.5cm
      - right=2.5cm
    papersize: a4
    mainfont: TeX Gyre Pagella
    mathfont: TeX Gyre Pagella Math
bibliography: papers.bib
---

# Overview

SIRCA’s end of day price service is a comprehensive collection of datasets designed to give you everything needed to find daily closing price for ASX listed companies, and to generate reliable daily measures of total shareholder returns.
Everything is retained for delisted companies.
Starting from January 2000, this collection includes all available end-of-day trade prices for the senior (residual risk) securities of all companies.
Dividend and franking histories; capital returns; adjustments for numerous corporate action events, like splits, consolidation, bonus issues, renounceable and non-renounceable issues; total daily traded volume and value, subdivided into various categories to expose activity in lit and dark ASX markets; numbers of issued shares; are all provided in this table.

All company names and ticker codes are recorded in a separate table, where these many pieces are linked together through time by an identifier we created for that purpose.
We call that identifier our Group Company Code, or `Gcode`.
`GCode` valuess are designed to make it easy for you to build price and return series for each company across its whole listed experience and not just that of pieces with different ticker codes.
Another goal for GCode is to remain unchanging across time.
`GCode` persistence means you can select a sample identified by `GCode`, test your ideas and come back later to update your research, without any extra work caused by intervening name and ticker code changes.
These same `GCodes` are used in many of our services, for the same beneficial reasons.

Other components of this service are our history of value- and equal-weighted all-of-market daily returns, which we generate from all observable daily company returns; another file identifying all of the ASX trading days since the start of January 2000, which can be used to identify gaps in company price series, from suspensions or thin trading; and a detailed description of all tables and fields in a data dictionary created just for this collection, which clearly describes how to interpret and use its many elements.

This entire collection is an ideal choice for researchers considering event studies for ASX listed companies.
With these data sets you can:

1.  Find `gcode`s from company names or ticker codes
2.  Apply the `cumulativefactor` field to adjust prices for different dividend and corporate action events.
3.  Generate and plot total shareholder returns
4.  Use the `dayssince` column to identify return intervals between consecutive trades
5.  Use the `seniorsecurity` column to restrict attention to the residual risk security of each company
6.  Recognise and accommodate negative factors and zero volumes
7.  Calculate cumulative factor without dividends
8.  Segment market trading activity by trade and venue type, such as to explore lit versus dark market trading.

# Tables

SIRCA's ASX end of day price collection includes the following tables:

| Table              | Description                                                                  | Primary key                                 |
|------------------------|------------------------------------------------------------------|-------------------|
| `si_au_ref_names`  | Complete name histories and change dates for companies listed from 01Jan2000 | `GCode`, ASX SecurityTicker, SeniorSecurity |
| `si_au_prc_daily`  | Complete daily price, volume and value histories, with issued share numbers  | `GCode`, ASX SecurityTicker, SeniorSecurity |
| `si_au_retn_mkt`   | Daily value- and equal-weighted whole market returns                         | `date`                                      |
| `si_au_ref_trdday` | Record of ASX trading dates since 01Jan2000                                  | `date`                                      |

# Statistical software examples

The following examples demonstrate ways in which you might extract content of interest to you, or work with data from more than one data set.
They are intended to help with analysis in your preferred environment, rather than to be precise guidance, because your favoured environment may require modifications to the exact form of these examples.

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(DBI)
library(dbplyr, warn.conflicts = FALSE)
library(duckdb)
library(farr)

conflicted::conflict_prefer("filter", "dplyr")
conflicted::conflict_prefer("lag", "dplyr")
```

```{r}
Sys.setenv(DATA_DIR = "~/Dropbox/pq_data")
```

```{r}
db <- dbConnect(duckdb::duckdb())

si_au_ref_names <-
  load_parquet(db, "si_au_ref_names", "sirca") |>
  compute(name = "si_au_ref_names")

si_au_prc_daily <- load_parquet(db, "si_au_prc_daily", "sirca")
si_au_ref_trddays <- load_parquet(db, "si_au_ref_trddays", "sirca")
```

## 1. Finding `GCodes` from company names or ticker codes

One way of searching for a `GCode` is to look up the company name, for example, every company name including Westpac, such as Westpac Banking Corporation, Westpac Office Trust, Westpac Property Trust.

```{sql}
#| connection: db
#| eval: false
SELECT gcode, securityticker, abbrevcompanyname
FROM si_au_ref_names
WHERE fullcompanyname LIKE '%WESTPAC%'
```

The same query could be run using `dplyr` code:

```{r}
si_au_ref_names |>
  filter(str_like(fullcompanyname, '%WESTPAC%')) |>
  select(gcode, securityticker, abbrevcompanyname) |>
  collect()
```

Alternatively, searching by the ticker code will also yield a result.

```{sql}
#| connection: db
#| eval: false
SELECT gcode, securityticker, abbrevcompanyname
FROM si_au_ref_names 
WHERE securityticker = 'ANZ'
```

Again the same query could be run using `dplyr` code:

```{r}
si_au_ref_names |>
  filter(securityticker == 'ANZ') |>
  select(gcode, securityticker, abbrevcompanyname) |>
  collect()
```

As another example, suppose one is interested in Arena REIT which has a CompanyTicker of `ARF`.
Searching for this ticker code reveals the `gcode` of Arena REIT is `arf2`.
This `gcode` can then be used to search the `si_au_prc_daily` table for information about the securities of Arena REIT.

```{r}
si_au_ref_names |>
  filter(companyticker == 'ARF') |>
  select(gcode, securityticker, abbrevcompanyname, 
         listdate, delistdate) |>
  arrange(listdate) |>
  collect()
```

The previous search reveals that in 2000, the ticker code ARF was then associated with Arrowfield Group Limited.
Arrowfield Group Limited is a different entity to Arena REIT, which was listed in 2013, so the two entities have separate `GCode`s.

A search for `CompanyTicker` equalling ARF also shows that in 2013, the `SecurityTicker` of Arena REIT briefly changed from `ARF` to `ARFDA`, and then back to `ARF`, due to conversions to and from deferred units.
Since Arena REIT remained the same entity during this process, the `GCode` remains `arf2` throughout this history, allowing for accurate comparisons across name changes and ticker code changes.

## 2. Adjusting for the effects of corporate actions

The `CumulativeFactor` column can be used to adjust the closing price for the effects of corporate actions, e.g. stock splits or entitlement offers, or dividends.

The following plot shows a large drop in the closing price of BHP in late June 2001.

```{r}
bhp <- 
  si_au_prc_daily |>
  filter(gcode == 'bhp1',
         between(date, '2001-01-01', '2001-12-31')) |>
  select(gcode, date, close, coraxdescription)
```

```{r}
bhp |>
  collect() |>
  ggplot(aes(x = date, y = close)) +
  geom_line()
```

Examining the `coraxdescription` column, it seems that the change is due to a `1:0.94` bonus issue.

```{r}
bhp |>
  filter(!is.na(coraxdescription)) |>
  collect()
```

The `coraxdescription` column provides details of CORAX (Corporate Action) events, when available.
However, descriptions are not always available.
The `numberofdilutionevents` field will always show a value greater than 0 when a dilution event (CORAX or dividend) has been considered and applied and `numberofcoraxevents > 0` specifically indicates CORAX events.
Likewise, `numberofdividendevents > 0` can be used to find all dividend events, even when other descriptive fields are not available.

```{r}
si_au_prc_daily |>
  filter(gcode == 'bhp1', 
          between(date, '2001-06-27', '2001-07-02')) |>
  select(gcode, date, close, numberofcoraxevents, coraxdescription) |>
  collect()
```

Simply multiplying the `close` column by the `cumulativefactor` column will produce the adjusted price.

```{r}
bhp <- 
  si_au_prc_daily |>
  filter(gcode == 'bhp1',
         between(date, '2001-01-01', '2001-12-31')) |>
  mutate(adjustedprice = close * cumulativefactor) |>
  select(gcode, date, close, adjustedprice, coraxdescription) |>
  collect()
```

In this case the new `adjustedprice` series is everywhere lower than `close` because `cumulativefactor` is affected by all CORAX events to the end of each series of Close prices.
This is so the resulting `adjustedprice` series is consistent over its entire history and can reliably measure returns between any two trading dates for `bhp1`.
Exactly the same process for `cumulativefactor` applies for dividends as well as corporate actions.
`AAA` is a security that pays regular dividends, and the effect of its dividends on the closing price can be observed from a graph.

```{r}
aaa <-
  si_au_prc_daily |>
  filter(gcode == 'aaa2', between(date, '2017-01-01', '2018-01-05')) |>
  mutate(adjustedprice = close * cumulativefactor) |>
  select(date, close, adjustedprice, cumulativefactor) |>
  collect()

aaa

aaa |>
  select(-cumulativefactor) |>
  pivot_longer(-date, names_to = "variable", values_to = "price") |>
  ggplot(aes(x = date, y = price, color = variable)) +
  geom_line()
```

## 3. Plotting a distribution of price relatives for a security.

The following function can calculate the price relative, or total shareholder return, for a security.
Note that it requires the price to be adjusted for Corporate Actions and Dividends, and to be ordered by Date.

```{r}
cba <- 
  si_au_prc_daily |>
  filter(gcode == 'cba1') |>
  mutate(adjustedprice = close * cumulativefactor) |>
  group_by(gcode) |>
  window_order(date) |>
  mutate(prel = adjustedprice / lag(adjustedprice)) |>
  ungroup() |>
  window_order() |>
  select(gcode, date, close, adjustedprice, prel, dayssince) |>
  compute()

cba |> collect(n = 10)
```

## 4. `dayssince` column

It is important to note that price relatives calculated in the previous section may not always be over consecutive trading days.
Although CBA is a stock that is consistently traded, a less-liquid security may show large gaps in trading activity, leading to price relatives that span larger time periods An example of this is shown below, using the `datesince` column.
This column contains the number of elapsed days since a point in time, and provides a way to measure the liquidity of a security.

```{r}
cba |>
  group_by(gcode) |>
  window_order(date) |>
  mutate(days_elapsed = dayssince - lag(dayssince)) |>
  count(days_elapsed, sort = TRUE)
```

Note we need to use `window_order()` in place of `arrange()` when using "remote" data frames.

```{r}
si_au_prc_daily |> 
  filter(gcode == '1st1') |>
  group_by(gcode) |>
  window_order(date) |>
  mutate(days_elapsed = dayssince - lag(dayssince)) |>
  ungroup() |>
  count(gcode, days_elapsed, sort = TRUE) |>
  collect()
```

```{r}
si_au_ref_trddays

si_au_prc_daily_dates <-
  si_au_prc_daily |>
  distinct(date) |>
  compute()

si_au_prc_daily_dates |>
  anti_join(si_au_ref_trddays, by = "date") |>
  inner_join(si_au_prc_daily, by = "date") |>
  count(date) |>
  mutate(wday = wday(date, label = TRUE)) |>
  arrange(desc(n)) |>
  collect()
```

```{r}
trading_days <- 
  si_au_ref_trddays |>
  window_order(date) |>
  mutate(td = row_number()) |>
  distinct(date, td) |>
  arrange(date) |>
  compute()
```

```{r}
days_elapsed_df <-
  si_au_prc_daily |>
  filter(gcode %in% c('cba1', '1st1')) |>
  inner_join(trading_days, by = "date") |>
  group_by(gcode) |>
  window_order(date) |>
  mutate(days_elapsed = td - lag(td),
         lag_date = lag(date)) |>
  select(gcode, date, lag_date, days_elapsed) |>
  ungroup() |>
  compute()
```

```{r}
days_elapsed_df |>
  count(gcode, days_elapsed) |>
  arrange(desc(gcode), desc(days_elapsed)) |>
  collect()
```

```{r}
days_elapsed_df |>
  filter(gcode == 'cba1', days_elapsed > 1)
```

It turns out there was a trading halt placed on [12 August 2015](https://announcements.asx.com.au/asxpdf/20150812/pdf/430fwrn8xk90lg.pdf).

## 5. `seniorsecurity` column

At times, some `gcodes` have multiple securities trading simultaneously.
This is because these companies have more than one class of security and a junior security class has been included in `si_au_prc_daily`.
Over time, other junior security classes may be included too, which means it is important to keep the different price series consistent.
The `seniorsecurity` column is included so you can focus on just the senior security series for any company.
In the next example, two classes of security are shown to be simultaneously trading for Telstra Corporation Ltd, whose `gcode` is `tls1`.
These are evident from the different `securityticker` values: `TLS` and `TLSCA`.
Notice the `seniorsecurity` field correctly separates these two series.

```{r}
si_au_prc_daily |>
  filter(gcode == 'tls1', between(date, "2008-05-01", "2008-05-31")) |>
  select(gcode, date, securityticker, seniorsecurity) |>
  arrange(date)
```

# 6. Negative factors and zero `volumeonmkt` values

The `factor` field may show a value of `-1` in some cases.
In order for a factor to be calculated for an event, there must be an observed trade price before the event and a price after the event.
When either of these prices does not exist, a factor of `-1` is assigned to the event.
This can occur both in the beginning and the end of the lifetime of the security.

The following example shows a dividend on `2004-07-05`.
However, no price was observed after the event, and hence the `factor` and `dividendfactor` fields contain a value of `-1`.
Note that there is a price in the `close` field on `2004-07-05` but it was not observed that day, after the dividend event.
This is evident from the 0 value for `VolumeOnMkt`, and confirmed by `NA` or `0` values for `open`, `high`, `low`.
This price is simply the previous observed trade price carried forward.

```{r}
si_au_prc_daily |> 
  filter(gcode == 'npx1', date <= '2002-03-18', seniorsecurity == 1) |>
  select(gcode, date, close, dividend, factor, dividendfactor, volumeonmkt) |>
  collect()
```

The following example shows dividends between `2000-03-06` and `2001-09-28` without any trading.
As no trading was observed prior to these dividend events, the `factor` and `dividendfactor` fields contain a value of `-1`.

```{r}
si_au_prc_daily |> 
 filter(gcode == 'wsf1', date >= '2004-07-01', seniorsecurity == 1) |>
  select(gcode, date, close, dividend, factor, dividendfactor, volumeonmkt) |>
  collect()
```

## 7. Calculating cumulative factor without including dividends

The provided `cumulativefactor` field is calculated by cumulating the `factor` column.
This adjustment includes both corporate actions and dividends and is included in the table for convenience.
For users wishing to calculate an adjustment without including dividends, the following example is provided for reference.
It uses the `adjustmentfactor` field, which provides dilution factors for just the CORAX events (when followed at some time by a valid `close` price).
Please note the following code converts all factors of `-1` into factors of `1`.
This is appropriate if the unobserved price does not change following the event leading to the factor of `-1`.
The function may not be appropriate for your particular application if that assumed pricing behaviour is invalid.

```{r}
si_au_prc_daily |>
  filter(adjustmentfactor < 0) |>
  count(adjustmentfactor)
```

```{r}
si_au_prc_daily |>
  group_by(gcode, seniorsecurity) |>
  window_order(desc(date)) |>
  mutate(cumulativefactor2 = exp(cumsum(log(abs(factor))))) |>
  window_order() |>
  ungroup() |>
  filter(gcode == "ikw1", between(date, "2018-11-20", "2018-11-30")) |>
  select(gcode, date, adjustmentfactor, cumulativefactor, close, cumulativefactor2) |>
  arrange(desc(date))
```

```{r}
si_au_prc_daily |>
  group_by(gcode, seniorsecurity) |>
  window_order(desc(date)) |>
  mutate(cumulativefactor2 = exp(cumsum(log(abs(factor))))) |>
  mutate(cumulativefactor2 = lag(cumulativefactor2) * sign(lag(factor))) |>
  window_order() |>
  ungroup() |>
  filter(gcode == "par1", between(date, "2018-12-03", "2019-02-18")) |> # par1
  select(gcode, date, adjustmentfactor, cumulativefactor, close, cumulativefactor2) |>
  arrange(desc(date))
```

Use the `adjustmentfactor` field, which does not account for dividends.
Visualise the new adjustment and compare to the adjustment from the example in part 1 Note that the CorporateOnlyAdjustedPrice, which is calculated without including dividends, looks identical to the Close price as no corporate actions have occurred within this time frame.

The following shows the effect that dividends can have on the adjusted price series.
The AdjustedPrice series incorporates both CORAX factors and dividend factors, whereas the CorpAdjustedPrice series incorporates only CORAX adjustments and ignores any dividends.
The CORAX-only price series shows a visible fall at the time when the dividend occurs as the value of the dividend is not accounted for.

```{r}
si_au_prc_daily |> 
 filter(gcode == 'dmg1', 
        between(date, "2012-07-01", "2012-07-31"), 
        seniorsecurity == 1) |>
  select(gcode, date, close, dividend, factor, dividendfactor, volumeonmkt) |>
  collect()
```

```{r}
#| eval: false
dmg <- sql("SELECT * FROM si_au_prc_daily WHERE GCode = '' AND Date BETWEEN '2012-07-01' AND '2012-07-31' AND SeniorSecurity = 1")
dmg <- collect(dmg)
dmg$CorpFactor <- cumulativeFactor(dmg$AdjustmentFactor) # Use the AdjustmentFactor field, covering corporate adjustments only (no dividends).
dmg$CorpAdjustedPrice <- dmg$CorpFactor * dmg$Close # Not accounting for dividends. dmg$AdjustedPrice <- dmg$CumulativeFactor * dmg$Close # Calculated using the provided CumulativeFactor field, which accounts for dividends.
display(dmg)
```

Note the large difference between the different forms of adjusted price.
As revealed by the command above, this is due to the dividend of 0.4497 on 2012-07-09, which precedes a fall in Close price from 0.575 to 0.019.
Adjusting only for CORAX events can clearly lead to significantly different measures of share price performance when dividends are also present.
CORAX adjustments may also be used to standardise earnings information through time.

# 8. Segmentation by Trade Type

The `si_au_prc_daily` table also contains information on the count/volume/value of trades by various trade type categories.
This following section will show examples of how trading activity can be aggregated by different trade types:

1.  Trading activity across the whole market.
2.  Segmentation by on vs off-market trades.
3.  Proportion of on-market non-crossing trades that are carried out through ASX Centre Point
4.  Comparison of lit-pool and dark-pool trading

A graph from the following command shows the value of trading activity across the year plotted against time.
Trading activity can vary significantly from month to month.

```{r}
#| eval: false
wholemkt <- sql(" SELECT
Date,
sum(ValueOnMKt + ValueOffMkt) AS ValueWholeMkt FROM si_au_prc_daily
WHERE Date BETWEEN '2016-01-01' AND '2019-12-31' GROUP BY Date
ORDER BY Date
") 
display(wholemkt)
```

```{r}
#| eval: false
onvsoff <- sql(" SELECT
Date,
sum(VolumeOnMkt) AS VolumeOnMkt, sum(VolumeOffMkt) AS VolumeOffMkt
FROM si_au_prc_daily
WHERE Date BETWEEN '2019-01-01' AND '2019-12-31' GROUP BY Date
ORDER BY Date")
display(onvsoff)
```

Another distinction in trading activity is CentrePoint trades.
The ASX Centre Point matching system is a market for dark pool liquidity (more information on dark trades in the next section).
As such, CentrePoint trades are a subset of On-market trades.
More information on ASX Centre Point can be found on the [ASX website](https://www.asx.com.au/markets/trade-our-cash-market/asx-equities-trading/asx-centre-point.html).
The composition of each market segment is displayed at the bottom of the `si_au_prc_daily` tab in our data dictionary for this service.
The following example shows the average proportion of on-market non- crossing trades that are directed through ASX Centre Point over a period of time.

```{r}
#| eval: false
centrepoint <- sql(" SELECT
Date,
avg((ValueCentrePt - ValueCentrePtCross) / (ValueOnMkt - ValueOnMktCross)) AS AvgPropCentrePtNonCross
FROM si_au_prc_daily
WHERE Date BETWEEN '2018-01-01' AND '2019-12-31'
AND ValueOnMkt > 0 AND ValueCentrePt > 0 GROUP BY Date
ORDER BY Date
")
display(centrepoint)
```

It is a simple matter to focus on particular market segments.
For example the previous query can be targeted on companies whose market capitalisation is less then \$50 million.

```{r}
#| eval: false
centrepoint_small <- sql(" SELECT
Date,
avg((ValueCentrePt - ValueCentrePtCross) / (ValueOnMkt - ValueOnMktCross)) AS AvgPropSmlCtPtNonCross
FROM si_au_prc_daily
WHERE Date BETWEEN '2018-01-01' AND '2019-12-31'
AND ValueOnMkt > 0 AND ValueCentrePt > 0 AND (Close * Shares) < 50000000
GROUP BY Date
ORDER BY Date
") 
display(centrepoint_small)
```

As mentioned above, it is possible to segment the market by the visibility of trades.
In the lit pool, or LIT market, the order book is public and all bid/offer orders are visible to all participants.
That is in contrast to the dark pool, or DARK market, where the order book is not visible until trades are executed.
The dark pool consists of both on-market and off-market crossing trades, as well as any CentrePoint trades as noted previously.
This following section shows the distribution of activity across the LIT/DARK markets over time.
Note: As our CentrePoint trade measures include crossing trades, CentrePoint crossing trade volumes need to be subtracted to avoid double-counting these trades in the calculation of the dark pool trading.

```{r}
#| eval: false
litvsdark <- sql(" SELECT
Date, 
sum(VolumeOffMktCross + VolumeOnMktCross + VolumeCentrePt - VolumeCentrePtCross) AS
VolumeDark,
sum(VolumeOnMKt + VolumeOffMkt - (VolumeOffMktCross + VolumeOnMktCross + VolumeCentrePt - VolumeCentrePtCross)) AS VolumeLit
FROM si_au_prc_daily
WHERE Date BETWEEN '2017-01-01' AND '2019-12-31'
AND VolumeOnMkt > 0 GROUP BY Date ORDER BY Date
")
display(litvsdark)
```
