---
title: "Converting lazy data frames into Parquet files"
author: Ian D. Gow
date: 2026-03-22
date-format: "D MMMM YYYY"
categories: [R, Parquet, db2pq, WRDS]
format:
  html:
    colorlinks: true
    number-sections: true
  pdf: 
    number-sections: true
    colorlinks: true
    geometry:
      - left=2cm
      - right=2cm
    papersize: a4
    filters:
      - abstract.lua
    include-in-header: preamble.tex
    mainfont: TeX Gyre Pagella
    mathfont: TeX Gyre Pagella Math
bibliography: papers.bib
csl: jfe.csl
execute:
  freeze: auto
keywords:
  - R
  - Arrow
  - DuckDB
---

::: {.abstract}
In this short note, I show how my `db2pq` package can be used to turn "lazy data frames" produced by `dbplyr` into Parquet files on disk without needing to load the data into RAM and with minimal RAM requirements.
:::

```{r}
#| include: false
Sys.setenv(PGHOST = "wrds-pgdata.wharton.upenn.edu",
           PGPORT = 9737L,
           PGDATABASE = "wrds",
           PGUSER = "iangow")
```

Need `pak::pak("iangow/db2pqr")` first

```{r}
#| warning: false
#| message: false
library(tidyverse)
library(DBI)
library(arrow)
library(farr)
library(db2pq)
library(wrds)
```

# Motivating example

While many of the ideas I discuss here are general in nature, for concreteness, I focus on the special case of a WRDS user maintaining a local Parquet data library of the kind discussed in [Appendix E](https://iangow.github.io/far_book/parquet-wrds.html#sec-make-pq) of [*Empirical Research in Accounting: Tools and Methods*](https://iangow.github.io/far_book/) and provide examples using my R package `db2pq`.

In 



```{r}
wrds <- wrds_connect()

funda <- tbl(wrds, Id(schema = "comp", table = "funda"))

compustat <- 
  funda |>
  filter(indfmt == 'INDL', datafmt == 'STD', 
         popsrc == 'D', consol == 'C') |>
  mutate(mkt_cap = prcc_f * csho,
         size = if_else(mkt_cap > 0, log(mkt_cap), NA),
         roa = if_else(at > 0, ib / at, NA),
         mtb = if_else(ceq > 0, mkt_cap / ceq, NA),
         special_items = if_else(at > 0, coalesce(spi, 0) / at, NA),
         fas133 = !is.na(aocidergl) & aocidergl != 0,
         fas157 = !is.na(tfva) & tfva != 0) |>
  select(gvkey, iid, datadate, mkt_cap, size, roa, mtb,
         special_items, fas133, fas157) |>
  filter(mkt_cap > 0) |>
  mutate(eomonth = floor_date(datadate, "month") + months(1) - days(1)) |>
  mutate(eomonth = as.Date(eomonth))
```

```{r}
compustat
```

To facilitate the creation of a data repository for the empirical example in that chapter, we created `save_parquet()`, which takes a (possibly remote) data frame and uses the `write_parquet()` function from the `arrow` library to save it as a Parquet file.^[The Parquet format is described in *R for Data Science* [@Wickham:2023aa, p. 393] as "an open standards-based format widely used by big data systems."
Parquet files provide a format optimized for data analysis, with a rich type system.
More details on the Parquet format can be found in [Chapter 22](https://r4ds.hadley.nz/arrow) of *R for Data Science*.]

```{r}
save_parquet <- function(df, name, schema = "", path = data_dir) {
  file_path <- file.path(path, schema, str_c(name, ".parquet"))
  write_parquet(collect(df), sink = file_path)
}
```

```{r}
data_dir <- Sys.getenv("DATA_DIR")

compustat_pq <-
    save_parquet(compustat, "compustat", "glms") |>
    system_time()
```


```{r}
save_parquet <- function(df, name, schema = "", path = data_dir) {
  file_path <- file.path(path, schema, str_c(name, ".parquet"))
  lazy_tbl_to_pq(df, out_file = file_path)
}
```

```{r}
compustat_path <-
    save_parquet(compustat, "compustat", "glms") |>
    system_time()
```

```{r}
compustat_pq <- open_dataset(compustat_path)
```

```{r}
compustat_pq |> 
    filter(gvkey == "001161") |>
    collect()
```

```{r}
db <- dbConnect(duckdb::duckdb())

compustat_db <- load_parquet(db, "compustat", "glms")
compustat_db |> 
    filter(gvkey == "001161") |>
    collect()

dbDisconnect(db)
```