---
title: "SIRCA Mergers and Acquisitions collection"
author: 
  - name: Ian D. Gow
    orcid: 0000-0002-6243-8409
    email: iandgow@gmail.com
date: 2025-09-11
categories: [Australia, "M&A", ASX, SIRCA, Parquet]
date-format: "D MMMM YYYY"
format:
  html:
    colorlinks: true
  pdf: 
    colorlinks: true
    geometry:
      - left=2.5cm
      - right=2.5cm
    papersize: a4
    mainfont: TeX Gyre Pagella
    mathfont: TeX Gyre Pagella Math
bibliography: papers.bib
execute:
  freeze: auto
---

# Overview

SIRCA's Mergers and Acquisitions collection provides data on Australian M&A transactions.

The code in this note uses the packages listed below, plus the `duckdb` package.^[Execute `install.packages(c("tidyverse", "arrow", "duckdb", "DBI"))` within R to install all the packages you need to run the code in this note.]
This note was written using [Quarto](https://quarto.org) and compiled with [RStudio](https://posit.co/products/open-source/rstudio/), an integrated development environment (IDE) for working with R.
The source code for this note is available [here](https://raw.githubusercontent.com/iangow/notes/main/sirca_ma.qmd).

```{r}
#| include: false
library(tinytable)

knit_print_alt <- function(x, ...) {
  res <- knitr::knit_print(knitr::kable(x, digits = 3))
  knitr::asis_output(res)
}
```

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(arrow)
```

# Getting SIRCA Merger and Acquisitions data

SIRCA supplies the data we will use as two compressed CSV files.
I provide the code needed to prepare the data for the analysis below.
I merely assume that you have access to SIRCA and have downloaded the raw data from SIRCA along the lines discussed below.^[See [SIRCA's documentation](http://www.sirca.org.au/wp-content/uploads/2023/11/How-to-use-the-Data-Library.pdf) for details on getting the data.]
Once you have done that you should be able to execute all the code in this note and thereby produce all the output (tables and graphs) contained herein.

You should download these files and place them *as is* in a single `sirca` directory and you should edit the code below to tell my script where to look for that directory.

In my case, `sirca` is found in `~/Dropbox/raw_data` and so I execute the following command in RStudio:

```{r}
Sys.setenv(RAW_DATA_DIR = "~/Dropbox/raw_data")
```

The script also needs to know where to put the processed data files, which will be in parquet format.
You should edit the following line to refer to a location on *your* computer that you would like to use to store processed SIRCA data.

```{r}
#| eval: false
Sys.setenv(DATA_DIR = "~/Dropbox/pq_data")
```

```{r}
pq_dir <- file.path(Sys.getenv("DATA_DIR"), "sirca")
ma_delisted_file <- file.path(Sys.getenv("RAW_DATA_DIR"), 
                              "sirca",
                              "Delisted_MergerAndAcquisition-2025-08.csv.gz")

ma_delisted <- read_csv(ma_delisted_file, show_col_types = FALSE)


ma_other_file <- file.path(Sys.getenv("RAW_DATA_DIR"), 
                              "sirca",
                              "MergerAndAcquisition-2025-08.csv.gz")

ma_other <- read_csv(ma_other_file, show_col_types = FALSE)

si_ms_ma <-
  ma_delisted |>
  union_all(ma_other) |> 
  write_parquet(sink = file.path(pq_dir, "si_ms_ma.parquet"))
```
