---
title: "Defining winter and summer in Melbourne"
author: 
  - name: Ian D. Gow^[University of Melbourne, [ian.gow@unimelb.edu.au](mailto:ian.gow@unimelb.edu.au)]
    orcid: 0000-0002-6243-8409
    email: iandgow@gmail.com
date: 2024-04-20
date-format: "D MMMM YYYY"
bibliography: papers.bib
title-block-style: plain
csl: jfe.csl
format: 
  pdf:
    toc: false
    number-sections: true
    colorlinks: true
    mainfont: TeX Gyre Pagella
    sansfont: TeX Gyre Pagella
  html:
    toc: false
    number-sections: true
    colorlinks: true
execute: 
  freeze: auto
keywords:
  - Python
---

```{r}
#| label: setup
#| include: false
Sys.setenv(RETICULATE_PYTHON = "../notes_py/bin/python")
library(reticulate)
py_config()
```

```{r}
#| include: false
library(tidyverse)
library(DBI)
library(dbplyr)
library(scales)
```

```{python}
#| label: py_get_data
#| include: false
import openmeteo_requests

import requests_cache
import pandas as pd
from retry_requests import retry

# Setup the Open-Meteo API client with cache and retry on error
cache_session = requests_cache.CachedSession('.cache', expire_after = -1)
retry_session = retry(cache_session, retries = 5, backoff_factor = 0.2)
openmeteo = openmeteo_requests.Client(session = retry_session)

latitude = -37.814
longitude = 144.96332
start_date = "2001-01-01"
end_date = "2023-12-31"
timezone = "Australia/Melbourne"

# Make sure all required weather variables are listed here
# The order of variables in hourly or daily is important to assign them correctly below
url = "https://archive-api.open-meteo.com/v1/archive"
params = {
	"latitude": latitude,
	"longitude": longitude,
	"start_date": start_date,
	"end_date": end_date,
	"hourly": "temperature_2m",
	"daily": ["temperature_2m_max", "temperature_2m_min"],
	"timezone": timezone
}
responses = openmeteo.weather_api(url, params=params)
response = responses[0]

# Process daily data. The order of variables needs to be the same as requested.
daily = response.Daily()
daily_temperature_max = daily.Variables(0).ValuesAsNumpy()
daily_temperature_min = daily.Variables(1).ValuesAsNumpy()

daily_data = {"date": pd.date_range(
	start = pd.to_datetime(daily.Time(), unit = "s", utc = True),
	end = pd.to_datetime(daily.TimeEnd(), unit = "s", utc = True),
	freq = pd.Timedelta(seconds = daily.Interval()),
	inclusive = "left"
)}
daily_data["temperature_2m_max"] = daily_temperature_max
daily_data["temperature_2m_min"] = daily_temperature_min

daily_dataframe = pd.DataFrame(data = daily_data)
```

```{r}
#| include: false
db <- dbConnect(duckdb::duckdb())

timezone = py$timezone
latitude = py$latitude
longitude = py$longitude

temps_daily <-
  py$daily_dataframe |>
  mutate(date = as.Date(date, tz = timezone))  |>
  copy_to(db, df = _, name = "temps_daily", overwrite = TRUE)
```

```{r}
#| include: false
season_length <- 91

mel_stats <-
  temps_daily |>
  window_order(date) |>
  window_frame(0, season_length - 1) |>
  mutate(n_days = n(),
         last_day = min(date, na.rm = TRUE),
         avg_min = mean(temperature_2m_min, na.rm = TRUE),
         avg_max = mean(temperature_2m_max, na.rm = TRUE)) |>
  arrange() |>
  compute()
```

```{r}
#| include: false
max_year <-
  mel_stats |>
  summarize(max(year(date))) |>
  pull()

max_n_days <-
  mel_stats |>
  summarize(max(n_days)) |>
  pull()

mel_avgs <-
  mel_stats |>
  ungroup() |>
  mutate(month = month(date), day = day(date), year = year(date)) |>
  filter(n_days == max_n_days, year < max_year) |>
  group_by(month, day) |>
  summarize(avg_min = mean(avg_min, na.rm = TRUE),
            avg_max = mean(avg_max, na.rm = TRUE),
            .groups = "drop") |>
  collect()
```

```{r}
#| include: false
mel_seasons <-
  mel_avgs |> 
  mutate(month = str_pad(month, width = 2, pad = "0"),
         day = str_pad(day, width = 2, pad = "0")) |>
  mutate(date = as.Date(str_c("2024", month, day, sep = "-"))) |>
  select(date, avg_min, avg_max) |>
  pivot_longer(cols = -date, names_to = "metric") |>
  mutate(metric = str_replace(metric, "^avg_", ""))
```

```{r}
#| include: false
winter_max <-
  mel_seasons |>
  filter(metric == "max") |>
  filter(value == min(value)) |>
  select(date) |>
  pull()

winter_min <-
  mel_seasons |>
  filter(metric == "min") |>
  filter(value == min(value)) |>
  select(date) |>
  pull()

summer_max <-
  mel_seasons |>
  filter(metric == "max") |>
  filter(value == max(value)) |>
  select(date) |>
  pull()

summer_min <-
  mel_seasons |>
  filter(metric == "min") |>
  filter(value == max(value)) |>
  select(date) |>
  pull()

hex <- hue_pal()(2)
```

```{r}
#| include: false
min_year <- mel_stats |> summarize(min(year(date))) |> pull()
max_year <- mel_stats |> summarize(max(year(date))) |> pull()

fig_cap <- str_c("Average daily temperatures for ", max_n_days, 
                 " days following indicated date for period ",
                 min_year, "--", max_year)
```

```{r}
#| label: fig-mel
#| fig-cap: !expr fig_cap
#| echo: false
mel_seasons |>
  ggplot(aes(x = date, y = value, color = metric)) + 
  geom_line() +
  geom_vline(xintercept = summer_max, color = hex[1]) +
  geom_vline(xintercept = summer_min, color = hex[2]) +
  geom_vline(xintercept = winter_max, color = hex[1]) +
  geom_vline(xintercept = winter_min, color = hex[2]) +
  scale_x_date(breaks = "1 month",
               labels = date_format("%b"),
               guide = guide_axis(angle = 90)) +
  theme(legend.position = "bottom")
```

In the United States, one often hears people speak of the "official" start of seasons.
Ironically, there seems to be nothing that is official about these dates.
However, there is consensus about the dates in the US.
The "official" start of summer is the summer solstice (for 2024: 21 December in Melbourne, 20 June in Boston) and the "official" start of winter is the winter solstice (for 2024: 21 June in Melbourne, 21 December in Boston).^[Seasons reckoned in this way are known as **astronomical seasons**. See [here](https://en.wikipedia.org/wiki/Season#Meteorological).]

In Australia, the usual convention is to divide seasons by months.
On this basis, winter starts on 1 June and summer starts on 1 December.^[Seasons reckoned in this way are known as **meteorological seasons**. See [here](https://en.wikipedia.org/wiki/Season#Astronomical).]

Is there a sense in which one approach is more correct than the other?
Focusing on summer and winter, one definition for these seasons would be that winter starts on the first day of the `r max_n_days`-day period that is the coldest such period for a year averaged over a number of years.
Similarly, summer should starts on the first day of the `r max_n_days`-day period that is the hottest such period for a year averaged over a number of years.

We answer this question focusing on Melbourne, Australia (latitude of `r latitude`, longitude: `r longitude`).

Daily temperature data from [Open-Meteo](https://open-meteo.com) comprise a maximum and minimum temperature.
So immediately we have two possible definitions of each season according to the temperature we use (e.g., summer could be the `r max_n_days`-day period that has the highest average minimum temperature or it could be the period that has the highest average maximum temperature).
Here we consider both.

The start of winter based on the `r max_n_days`-day period with the lowest average maximum temperature is **`r format(winter_max, "%d %B")`**.
The start of winter based on the `r max_n_days`-day period with the lowest average minimum temperature is **`r format(winter_min, "%d %B")`**.

The start of summer based on the `r max_n_days`-day period with the highest average maximum temperature is **`r format(summer_max, "%d %B")`**.
The start of summer based on the `r max_n_days`-day period with the highest average minimum temperature is **`r format(summer_min, "%d %B")`**.
So using maximums, we get close to the Australian convention for winter and close to the US convention for summer.


```{r}
#| include: false
spring_max <- winter_max + days(max_n_days)
autumn_max <- summer_max + days(max_n_days)
```

Interestingly, it seems that using average maximums for summer and winter gets closest to the current approach in Australia.
However, even using these we have the issue that spring begins on `r format(spring_max, "%d %B")` and autumn begins on `r format(autumn_max, "%d %B")`.
This implies a spring of `r summer_max - spring_max` days and an autumn of `r winter_max + years(1) - autumn_max` days.

```{r}
#| include: false
dbDisconnect(db)
```
